### 2023-07-12 manuscript/manuscript.qmd
initiated manuscript repository

---
title: "smdi: An R package to perform structural missing data investigations"

code-fold: false
eval: false
echo: true
number-sections: false

format: 
  docx:
    reference-doc: custom-reference-doc.docx
    fig-cap-location: top

editor: visual
#bibliography: references.bib
#csl: jamia.csl
---

```{r setup}
#| label: setup
#| include: false
#| eval: true
library(magick)
library(cowplot)
library(here)
```

**Authors**: Janick Weberpals^1^, Sudha R. Raman^2^, Pamela A. Shaw^3^, Hana Lee^4^, Massimiliano Russo^1^, Bradley G. Hammill^2^, Sengwee Toh^5^, John G. Connolly^5^, Kimberly J. Dandreo^5^, Fang Tian^4^, Wei Liu^4^, Jie Li^4^, José J. Hernández-Muñoz^4^, Robert J. Glynn^1^, Rishi J. Desai^1^

[Author affiliations:]{.underline}

^1^ Division of Pharmacoepidemiology and Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA

^2^ Department of Population Health Sciences, Duke University School of Medicine, Durham, NC

^3^ Biostatistics Division, Kaiser Permanente Washington Health Research Institute, Seattle, WA

^4^ Center for Drug Evaluation and Research, Food and Drug Administration, Silver Spring, MD

^5^ Department of Population Medicine, Harvard Medical School and Harvard Pilgrim Health Care Institute, Boston, MA

[**Correspondence:**]{.underline}

Janick Weberpals, RPh, PhD

Division of Pharmacoepidemiology and Pharmacoeconomics,

Department of Medicine, Brigham and Women's Hospital, Harvard Medical School,

1620 Tremont Street, Suite 3030-R, Boston, MA 02120, USA

Phone: 617-278-0932

Fax: 617-232-8602

Email: [jweberpals\@bwh.harvard.edu](mailto:jweberpals@bwh.harvard.edu)

[**Word count:**]{.underline} xxxx words

[**Tables:**]{.underline}

[**Figures:**]{.underline}

[**Supplementary material:**]{.underline}

[**Short running title**]{.underline}: An R package to perform structural missing data investigations

[**Keywords:**]{.underline} Missing data, Confounder, EHR, R, Software,

[**Funding:**]{.underline} This project was supported by Master Agreement 75F40119D10037 from the US Food and Drug Administration (FDA).

[**Disclosures/COI:res/COI:**]{.underline} The FDA approved the study protocol, statistical analysis plan and reviewed and approved this manuscript. Coauthors from the FDA participated in the results interpretation and in the preparation and decision to submit the manuscript for publication. The FDA had no role in data collection, management, or analysis. The views expressed are those of the authors and not necessarily those of the US FDA. Janick Weberpals reports prior employment by Hoffmann-La Roche and previously held shares in Hoffmann-La Roche. Pamela Shaw is a named inventor on a patent licensed to Novartis by the University of Pennsylvania for an unrelated project. Sengwee Toh serves as a consultant for Pfizer, Inc. and TriNetX, LLC.. Robert J Glynn has received research funding through his employer from Amarin, Kowa, Novartis, and Pfizer. Dr. Desai reports serving as Principal Investigator on investigator-initiated grants to the Brigham and Women's Hospital from Novartis, Vertex, and Bristol-Myers-Squibb on unrelated projects. All remaining authors report no disclosures or conflicts of interest.

[Analytical code and data sharing statement:]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2. All R code and materials can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>. Detailed information on packages and versions can be found in the `renv.lock` file in the linked repository. The R package and data used in this study can be downloaded from <https://janickweberpals.gitlab-pages.partners.org/smdi/>.

[**Acknowledgments:**]{.underline} None.

{{< pagebreak >}}

# Abstract {.unnumbered}

{{< pagebreak >}}

::: {.callout-note icon="false"}
## Highlights

-   ...
:::

{{< pagebreak >}}

# Introduction

{{< pagebreak >}}

# Methods

{{< pagebreak >}}

# Results

{{< pagebreak >}}

# Discussion

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Figures {.unnumbered}

```{r}
#| label: fig-RStudio-gui
#| fig-cap: "Steps to clone a remote repository using the RStudio graphical user interface."
#| eval: false
#| echo: false
#| out-width: 50%
knitr::include_graphics(here::here("figures", ""))
```

{{< pagebreak >}}

### 2023-07-12 manuscript/supplement.qmd
initiated manuscript repository

---
title: "Supplementary Appendix"
subtitle: "smdi: An R package to perform structural missing data investigations"
echo: false
number-sections: true

format: 
  pdf:
    fig-cap-location: top
    fig-align: center
    crossref:
      fig-title: "Supplementary Figure"
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true

editor: visual
#bibliography: references.bib
#csl: jamia.csl
---

```{r setup}
#| label: setup
#| include: false
#| eval: true
library(magick)
library(cowplot)
library(here)
library(knitr)
```

{{< pagebreak >}}

# Supplementary material

{{< pagebreak >}}

# Supplementary Figures

```{r}
#| label: fig-GitHub-gui_a
#| fig-cap: "Starting a new repository through the GitHub GUI."
#| out-width: 115%

include_graphics(here("figures", ""))
```

{{< pagebreak >}}

```{r}
#| label: fig-GitHub-gui_changes
#| eval: false
#| fig-cap: "Remote repositories like GitHub provide tools to visually track changes made to analytical code and corresponding output such as figures. This commit history illustrates details about the commit (e.g., commit SHA hash and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom)."
#| out-width: 105%

include_graphics(here("figures", "SFigure_4_commit_Github_gui.png"))
```

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

#  {.unnumbered}

### 2023-07-19 manuscript/manuscript.qmd
started manuscript and added table 1 (function overview)

@@
 ---
-title: "smdi: An R package to perform structural missing data investigations"
+title: "smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world databases"
 
@@
     fig-cap-location: top
+    tbl-cap-location: top
 
 editor: visual
-#bibliography: references.bib
-#csl: jamia.csl
+bibliography: references.bib
+csl: computer-methods-and-programs-in-biomedicine
+
+filters:
+  - docx-landscape.lua
 ---
@@
 library(here)
+library(tidyverse)
+library(readxl)
+library(gt)
+
+biblio <- bibtex::read.bib("references.bib")
 ```
@@
 Raman^2^, Pamela A.
-Shaw^3^, Hana Lee^4^, Massimiliano Russo^1^, Bradley G.
+Shaw^3^, Hana Lee^4^, Bradley G.
 Hammill^2^, Sengwee Toh^5^, John G.
@@
 
-^1^ Division of Pharmacoepidemiology and Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA
+^1^Division of Pharmacoepidemiology and Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA
 
-^2^ Department of Population Health Sciences, Duke University School of Medicine, Durham, NC
+^2^Department of Population Health Sciences, Duke University School of Medicine, Durham, NC
 
-^3^ Biostatistics Division, Kaiser Permanente Washington Health Research Institute, Seattle, WA
+^3^Biostatistics Division, Kaiser Permanente Washington Health Research Institute, Seattle, WA
 
-^4^ Center for Drug Evaluation and Research, Food and Drug Administration, Silver Spring, MD
+^4^Center for Drug Evaluation and Research, Food and Drug Administration, Silver Spring, MD
 
-^5^ Department of Population Medicine, Harvard Medical School and Harvard Pilgrim Health Care Institute, Boston, MA
+^5^Department of Population Medicine, Harvard Medical School and Harvard Pilgrim Health Care Institute, Boston, MA
 
@@
 
-[Analytical code and data sharing statement:]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2.
+[**Analytical code and data sharing statement:**]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2.
 All R code and materials can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>.
 Detailed information on packages and versions can be found in the `renv.lock` file in the linked repository.
-The R package and data used in this study can be downloaded from <https://janickweberpals.gitlab-pages.partners.org/smdi/>.
+The R package and data presented in this study can be downloaded from <https://janickweberpals.gitlab-pages.partners.org/smdi/>.
 
-[**Acknowledgments:**]{.underline} None.
+[**Acknowledgments:**]{.underline} We would like to thank all beta testers and attendees of the Division of Pharmacoepidemiology and Pharmacoeconomics Methods Incubator who gave valuable feedback on early versions of the `smdi` R package.
 
@@
 
-# Introduction
+# Background
+
+Electronic health records (EHR) are increasingly used to conduct real-world evidence (RWE) studies to complement evidence coming from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
+Due to their detailed capture of clinical parameters such as vital signs, lab measurements, physician assessments and lifestyle factors, EHRs can significantly improve the ability to control for confounding and imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
+However, such prognostic factors are often just partially observed which challenges the statistical analysis of the data and can result in severe bias when predicting or estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
+
+In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019].
+Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
+These methods comprise standard procedures such as comparing distributions of baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023; @hotelling1931; @little1988], checking the ability to predict missingness based on observed data [@sondhi2023] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] and guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
+However, these methods have so far been only described and tested in isolation from each other and no principled approach exists.
+In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed in both RWE and RCTs [@carroll2020; @wood2004; @harel2012].
+
+To overcome these shortcomings, we [@SentinelCI3] have recently developed and evaluated a principled approach combining a range of missing data diagnostics in an US EHR-claims database linkage [@weberpals2023].
+The results of this study revealed that the combination of these diagnostics characterized simulated missing data mechanisms well and provided helpful insights for the appropriate choice of analytic methods to handle the partially observed confounder data (e.g., missing data imputation).
+
+To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we here present and demonstrate the use of the `smdi` (structural missing data investigations) R package [@smdi].
+To that end, we first provide some brief background on the theoretical assumptions that underlie the diagnostic functions of this package.
+We then give an overview of the implemented functions and demonstrate an end-to-end workflow application of the package with a hypothetical RWE study using simulated oncology EHR dataset that comes bundled with the package.
 
@@
 # Methods
+
+## Problem formalization
+
+As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data (RWD) are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., EHR).
+Hence, confounders and prognostic factors, which need to be balanced between treatment groups in RWE studies are usually not available for all patients at all neccessary time points [@toh2012a].
+If the underlying mechanism for the missingness of such covariates is informative towards the outcome under study, e.g., patients who have a certain biomarker measured have a significantly increased risk if experiencing the outcome, this can lead to bias in the resulting effect estimates for the studied treatments under common missing data approaches like complete case analysis or imputation [@groenwold2020].
+Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for specific missing data approaches hold [@pedersen2017; @lee2013].
+
+## Theoretical background
+
+In our previous work [@weberpals2023] and for the purpose of the `smdi` R package, we categorize the implemented principled missing data investigations into three groups of diagnostics, which will be discussed in this section @tbl-overview
+
+Functions overview
+
+....
+
+Data generation for hypothetical study
+
+....
 
@@
 
+::: landscape
+# Tables {.unnumbered}
+
+```{r}
+#| label: tbl-overview
+#| tbl-cap: "table caption test"
+#| eval: true
+#| echo: false
+tbl1 <- readxl::read_excel(
+  path = here("tables", "tables_manual.xlsx"),
+  sheet = "table1"
+  ) %>% 
+  group_by(Group, `Group Description`) %>% 
+  gt() %>% 
+  tab_header(
+    title = "Overview of the main functions of the smdi package."
+    ) %>%
+  tab_caption(
+    caption = "Overview of the main functions of the smdi package."
+    ) %>%
+  tab_style(
+    cell_text(weight = "bold"),
+    locations = cells_column_labels(columns = everything())
+  ) %>% 
+  # tab_footnote(
+  #   footnote = "hotelling1931",
+  #   locations = cells_body(
+  #     columns = Function,
+  #     rows = Function == "**smdi_hotelling()**"
+  #     )
+  #   ) %>% 
+  fmt_markdown()
+
+tbl1
+```
+:::
+
+{{< pagebreak >}}
+
 # Figures {.unnumbered}

### 2023-07-19 manuscript/manuscript.qmd
call it a day

@@
 code-fold: false
-eval: false
 echo: true
@@
     fig-cap-location: top
-    tbl-cap-location: top
 
@@
 #| include: false
-#| eval: true
 library(magick)
@@
 library(gt)
+library(captioner)
 
-biblio <- bibtex::read.bib("references.bib")
+table_cap <- captioner(prefix = "Table")
 ```
@@
 
-In our previous work [@weberpals2023] and for the purpose of the `smdi` R package, we categorize the implemented principled missing data investigations into three groups of diagnostics, which will be discussed in this section @tbl-overview
+In our previous work [@weberpals2023] and for the purpose of the `smdi` R package, we categorized the implemented principled missing data investigations into three groups of diagnostics, which will be discussed in this section (Table 1).
 
@@
 ```{r}
-#| label: tbl-overview
-#| tbl-cap: "table caption test"
+#| label: tbl_overview
 #| eval: true
 #| echo: false
+
 tbl1 <- readxl::read_excel(
@@
   gt() %>% 
+  tab_caption(caption = "tbl-overview") %>% 
   tab_header(
-    title = "Overview of the main functions of the smdi package."
-    ) %>%
-  tab_caption(
-    caption = "Overview of the main functions of the smdi package."
+    title = "Overview of the main functions in smdi."
     ) %>%
@@
     ) %>% 
-  # tab_footnote(
-  #   footnote = "hotelling1931",
-  #   locations = cells_body(
-  #     columns = Function,
-  #     rows = Function == "**smdi_hotelling()**"
-  #     )
-  #   ) %>% 
   fmt_markdown()

### 2023-07-20 manuscript/manuscript.qmd
added section describing functions

@@
 library(gt)
-library(captioner)
-
-table_cap <- captioner(prefix = "Table")
 ```
@@
 Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
-These methods comprise standard procedures such as comparing distributions of baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023; @hotelling1931; @little1988], checking the ability to predict missingness based on observed data [@sondhi2023] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] and guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
+These methods comprise standard procedures such as comparing distributions of baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
 However, these methods have so far been only described and tested in isolation from each other and no principled approach exists.
@@
 To that end, we first provide some brief background on the theoretical assumptions that underlie the diagnostic functions of this package.
-We then give an overview of the implemented functions and demonstrate an end-to-end workflow application of the package with a hypothetical RWE study using simulated oncology EHR dataset that comes bundled with the package.
+We then give an overview of the implemented functions and demonstrate an end-to-end workflow application of the package with a hypothetical RWE study using a simulated oncology EHR dataset that comes bundled with the package.
 
@@
 As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data (RWD) are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., EHR).
-Hence, confounders and prognostic factors, which need to be balanced between treatment groups in RWE studies are usually not available for all patients at all neccessary time points [@toh2012a].
-If the underlying mechanism for the missingness of such covariates is informative towards the outcome under study, e.g., patients who have a certain biomarker measured have a significantly increased risk if experiencing the outcome, this can lead to bias in the resulting effect estimates for the studied treatments under common missing data approaches like complete case analysis or imputation [@groenwold2020].
+Hence, confounders and prognostic factors, which need to be balanced between treatment groups in RWE studies, are usually not available for all patients at all necessary time points [@toh2012a].
+If the underlying mechanism for the missingness of such covariates is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to be missing, this can lead to bias in the resulting effect estimates for the studied treatments under common missing data approaches like complete case analysis or imputation [@groenwold2020].
 Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for specific missing data approaches hold [@pedersen2017; @lee2013].
 
-## Theoretical background
+## Theoretical background and main package functions
 
-In our previous work [@weberpals2023] and for the purpose of the `smdi` R package, we categorized the implemented principled missing data investigations into three groups of diagnostics, which will be discussed in this section (Table 1).
+For the implementation of these routine diagnostics checks, we categorized the main functions of the `smdi` R package into three group diagnostics based on their general analytic purpose (Table 1).
+In this section, we will explain the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and how the results can be interpreted.
+Generally, for all functions in the `smdi` package, a *dataframe* is expected as input with a format where one row stands for one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates (one-row-per-patient format).
+Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the function.
+Throughout all functions, the user has the option to specify a vector with the column names of the partially observed covariate(s) that should be investigated (`covar` parameter).
+If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.
 
-Functions overview
+### Group 1 diagnostics
 
-....
+The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the partially observed covariate.
+According to Rubin's framework on inference and missing data [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates.
+Consequently, significant differences in patient characteristics would be expected between strata of patients with and without the partially observed covariate.
+If the missingness depends only on unobserved factors (missing not at random \[MNAR\]) or does not depend on either observed or unobserved covariates (missing completely at random \[MCAR\]), differences should not be observable.
 
-Data generation for hypothetical study
+To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics in the dataset [@schober2019; @austin2011; @tableone].
+To call the function, a user must specify the dataset th
+
+function returns an *amsd* object which displays an aggregated summary of the average or median (depending on the analyst's choice) and the minimum/maximum range of the derived ASMDs.
+The *amsd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate.
+For a graphical visualization, the function also creates a *ggplot2* graph [@ggplot2] illustrating the ASMDs for each compared patient characteristic in descending order.
+
+The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test.
+Hotelling's test [@hotelling1931; @Hotelling] formalizes this is a multivariate t-test for each partially observed covariate, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each partially observed covariate individually.
+In contrast, Little's test [@little1988; @naniar] (`smdi_little()`) computes a single global chi-square test statistic and p-value across all partially observed covariates with the null hypothesis that the data is MCAR.
+
+### Group 2 diagnostics
+
+The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` functions.
+This function trains and fits and random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors.
+If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism.
+In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism.
+The function returns an object of class `rf` which generically displays an overview of the AUC values of all partially observed covariates.
+For more details, the `rf` object additionally includes a *ggplot2* graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy.
+This metric can be valuable for interpreting and identifying strong predictors of missingness.
+It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model.
+For tuning the random forest model, the `smdi_rf()` function has a few parameters that can be specified by the user for optimization such as the number of trees to grow (default is 1,000 trees), the ratio of the size between train and test datasets (default is a 70/30 split) and the number of cores to parallelize the computation on [@parallel], since this function may be very time consuming, especially with larger datasets.
+
+### Group 3 diagnostics
+
+Finally, the third group diagnostics with the `smdi_outcome()` function examines association of the missingness indicator of the partially observed covariate and the outcome under study.
+The function will compute both a univariate model and a model adjusted for all other covariates specified in the dataset.
+Currently, three outcome regression types are supported: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes.
+Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be displayed exponentiated or not.
+
+## Data generation for hypothetical study
 
@@
 
-# Results
+# Results  {#sec-results}
 

### 2023-07-20 manuscript/manuscript.qmd
first draft of function description chapter

@@
 
-For the implementation of these routine diagnostics checks, we categorized the main functions of the `smdi` R package into three group diagnostics based on their general analytic purpose (Table 1).
-In this section, we will explain the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and how the results can be interpreted.
-Generally, for all functions in the `smdi` package, a *dataframe* is expected as input with a format where one row stands for one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates (one-row-per-patient format).
+For the implementation of these routine missing data diagnostics, we categorized the main functions of the `smdi` R package into three group diagnostics based on their general analytic prroperties (Table 1).
+In this section, we will explain the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and the interpretation of results.
+Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates (one-row-per-patient format).
 Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the function.
-Throughout all functions, the user has the option to specify a vector with the column names of the partially observed covariate(s) that should be investigated (`covar` parameter).
+Throughout all functions, the user has the option to specify a vector with the column name(s) of the partially observed covariate(s) that should be investigated (`covar` parameter).
 If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.
@@
 According to Rubin's framework on inference and missing data [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates.
-Consequently, significant differences in patient characteristics would be expected between strata of patients with and without the partially observed covariate.
+Consequently, significant differences in patient characteristics would be expected under a MAR mechanism between strata of patients with and without the partially observed covariate.
 If the missingness depends only on unobserved factors (missing not at random \[MNAR\]) or does not depend on either observed or unobserved covariates (missing completely at random \[MCAR\]), differences should not be observable.
@@
 To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics in the dataset [@schober2019; @austin2011; @tableone].
-To call the function, a user must specify the dataset th
-
-function returns an *amsd* object which displays an aggregated summary of the average or median (depending on the analyst's choice) and the minimum/maximum range of the derived ASMDs.
-The *amsd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate.
-For a graphical visualization, the function also creates a *ggplot2* graph [@ggplot2] illustrating the ASMDs for each compared patient characteristic in descending order.
+The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs (which can be flexibly chosen using the `median` parameter) along with the corresponding minimum-maximum ASMD range.
+The *asmd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate.
+For a graphical visualization of this, the function also creates a *ggplot2* graph [@ggplot2] illustrating the ASMDs for each compared patient characteristic in descending order.
 
@@
 The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` functions.
-This function trains and fits and random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors.
+This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors.
 If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism.
-In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism.
+In case of values close 0.5, this would rather indicate a random prediction and translate to potential MCAR or MNAR mechanisms.
+
 The function returns an object of class `rf` which generically displays an overview of the AUC values of all partially observed covariates.
+The AUC values are based on the prediction made in the respective test datasets.
 For more details, the `rf` object additionally includes a *ggplot2* graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy.
@@
 It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model.
-For tuning the random forest model, the `smdi_rf()` function has a few parameters that can be specified by the user for optimization such as the number of trees to grow (default is 1,000 trees), the ratio of the size between train and test datasets (default is a 70/30 split) and the number of cores to parallelize the computation on [@parallel], since this function may be very time consuming, especially with larger datasets.
+For tuning the random forest model, the `smdi_rf()` function has a few parameters that can be specified by the user for optimization such as the number of trees to grow (default is 1,000 trees), the ratio of the split between train and test datasets (default is a 70/30 split) and the number of cores to parallelize the computation on [@parallel], since this function may be very time consuming, especially with larger datasets.
 
@@
 Finally, the third group diagnostics with the `smdi_outcome()` function examines association of the missingness indicator of the partially observed covariate and the outcome under study.
-The function will compute both a univariate model and a model adjusted for all other covariates specified in the dataset.
-Currently, three outcome regression types are supported: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes.
+The function will compute both a univariate model and a model adjusted for all other covariates included in the dataset.
+In previous simulations, we observed characteristic patterns for univariate and adjusted associations of the missing indicator and the outcome that matched simulated underlying missingness mechanisms [@weberpals2023].
+As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the partially observed covariate.
+Under MAR, given that missingness can be explained by observed covariates, a potential spurious association in the univariate model disappeared after adjustment.
+If the missingness followed any MNAR mechanism, an association was observed in both univariate and adjusted models.
+
+Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes.
 Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be displayed exponentiated or not.
+The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each partially observed covariate.
 

### 2023-07-20 manuscript/manuscript.qmd
started with simulation cohort

@@
 #| include: false
-library(magick)
 library(cowplot)
@@
 library(gt)
+library(gtsummary)
+library(smdi)
 ```
@@
 To overcome these shortcomings, we [@SentinelCI3] have recently developed and evaluated a principled approach combining a range of missing data diagnostics in an US EHR-claims database linkage [@weberpals2023].
-The results of this study revealed that the combination of these diagnostics characterized simulated missing data mechanisms well and provided helpful insights for the appropriate choice of analytic methods to handle the partially observed confounder data (e.g., missing data imputation).
+The results of this large-scale study revealed that the combination of these diagnostics characterized simulated missing data mechanisms well and provided helpful insights for the appropriate choice of analytic methods to handle the partially observed confounder data (e.g., missing data imputation).
 
 To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we here present and demonstrate the use of the `smdi` (structural missing data investigations) R package [@smdi].
-To that end, we first provide some brief background on the theoretical assumptions that underlie the diagnostic functions of this package.
-We then give an overview of the implemented functions and demonstrate an end-to-end workflow application of the package with a hypothetical RWE study using a simulated oncology EHR dataset that comes bundled with the package.
+To that end, we first provide an overview on the package's main functions including the underlying theoretical assumptions.
+We then demonstrate the practical application of the package by illustrating and end-to-end workflow with a hypothetical RWE study using a simulated oncology EHR dataset that comes bundled with the package.
 
@@
 
-## Data generation for hypothetical study
+## Exemplary RWE study and data generation
 
-....
+To showcase the practical implementation of the functions provided in `smdi`, the package comes with simulated example EHR dataset focusing on an exemplary RWE study in the field of oncology with `r formatC(nrow(smdi_data), format = "fg", big.mark = ",")` lung cancer patients (Table 2).
+
+For this exemplary RWE study, we compared two systemic antineoplastic durg treatment regimens (`Drug A` versus `Drug B`).
+The true outcome of interest was overall survival (OS) which was defined as the time from treatment initiation to death due to any reason or censoring with a truncated follow-up time of 5 years (Supplementary Table and Supplementary Figure ).
+To simula The desired treatment effect estimate was defined with a hazard ratio (HR) of 1.0 and we assumed that the proportional hazards assumption holds.
+
+### Covariates
 
@@
 ```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl_cohort
+#| eval: true
+#| echo: false
+#| message: false
+
+labels <- readxl::read_excel(
+  path = here("tables", "tables_manual.xlsx"),
+  sheet = "covariates"
+  )
+
+labels_list <- as.list(labels$`Variable Label`)
+names(labels_list) <- labels$`Variable Name`
+
+tbl_cohort <- smdi_data %>% 
+  select(-c(eventtime, status)) %>% 
+  mutate(exposure = ifelse(exposure == 0, "Drug A", "Drug B")) %>% 
+  mutate(across(c(female_cat, smoking_cat, physical_cat, alk_cat, histology_cat, copd_cat), ~ifelse(.x == "0", 0, 1))) %>% 
+  tbl_summary(
+    by = "exposure", 
+    label = labels_list, 
+    missing = "ifany"
+    ) %>% 
+  add_overall() %>% 
+  add_difference()
+
+tbl_cohort
+```
 :::

### 2023-07-20 manuscript/supplement.qmd
started with simulation cohort

@@
 title: "Supplementary Appendix"
-subtitle: "smdi: An R package to perform structural missing data investigations"
-echo: false
+subtitle: "smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world databases"
+echo: true
 number-sections: true
@@
       fig-title: "Supplementary Figure"
+      tbl-title: "Supplementary Table"
     toc: true
@@
 #| eval: true
-library(magick)
-library(cowplot)
 library(here)
 library(knitr)
+library(tidyverse)
+library(survminer)
+library(gtsummary)
+library(readxl)
+
+seed_value <- 42
+n <- 2500
 ```
@@
 
-# Supplementary material
+# Supplementary Methods
+
+## Data Generation
+
+To give users the opportunity to get acquainted with the functionality and usage of the `smdi` package, the package includes a simulated example dataset.
+The dataset is an exemplary low-dimensional electronic health records (EHR) dataset depicting a cohort of `r formatC(n, format = "fg", big.mark = ",")` lung cancer patients.
+The dataset follows the general *one-row-per-patient* structure, in which one row stands for an individual patient and the columns represent the variables.
+
+### Exposure and outcome
+
+In this exemplary real-world evidence (RWE) study, the aim is to study the comparative effectiveness of two antineoplastic systemic drug treatment regimens (`Drug A versus Drug B`) on the time to all-cause mortalty due to any reason as the outcome (`overall survival`).
+The anticipated time of follow-up is truncated to 5 years.
+
+The desired strength of effectiveness of the **exposure of interest** is defined with a hazard ratio (HR) of 1.0, i.e.
+there is no difference in overall survival among patients who are treated with the exposure of interest as compared to the comparator regimen.
+The proportional hazards assumption is fulfilled for this dataset.
+
+### Confounders
+
+We further assume that there are some true confounders, which we need to specify to estimate our outcome model.
+Most of the covariates are associated with both the probability of treatment initiation and the outcome but there are also some that are not predictive of the exposure and just the outcome or not associated with any of the exposure or outcome, whatsoever.
+
+Despite the low dimensionality, the dataset is simulated as realistically as possible with varying strengths of associations between covariates and treatment initiation and the outcome.
+
+### Missingness
+
+Most importantly for testing the functionality of this package, some of the above mentioned confounders are just partially observed according to the missingness mechanisms and proportions specified below in the table below.
+
+### Overview covariates/confounder structure {#overview}
+
+To get an overview of the dataset, this table provides a summary of the different covariate-exposure-outcome-missingness correlations.
+
+```{r}
+read_excel(
+  path = here('tables', 'tables_manual.xlsx'),
+  sheet = "covariates"
+  ) %>% 
+  gt() %>% 
+  tab_header(
+    title = "Overview of covariates in simulated example dataset."
+    ) %>%
+  tab_style(
+    cell_text(weight = "bold"),
+    locations = cells_column_labels(columns = everything())
+    )
+```
+
+## Simulation of covariates and exposure
+
+```{r, covar_generation}
+# some global simulation parameters
+seed_value <- 42
+n <- 2500
+
+# start with basic dataframe, covariates and their association with exposure
+set.seed(seed_value)
+
+sim_covar <- tibble::tibble(
+  exposure = rbinom(n = n, size = 1, prob = 0.4),
+  age_num = rnorm(n, mean = 64 - 7.5*exposure, sd = 13.7),
+  female_cat = rbinom(n, size = 1, prob = 0.39 - 0.05*exposure),
+  ecog_cat = rbinom(n, size = 1, prob = 0.63 - 0.04*exposure),
+  smoking_cat = rbinom(n, size = 1, prob = 0.45 + 0.1*exposure), 
+  physical_cat = rbinom(n, size = 1, prob = 0.35 + 0.02*exposure),
+  egfr_cat = rbinom(n, size = 1, prob = 0.20 + 0.07*exposure), 
+  alk_cat = rbinom(n, size = 1, prob = 0.03),
+  pdl1_num = rnorm(n, mean = 40 + 10*exposure, sd = 10.5),
+  histology_cat = rbinom(n, size = 1, prob = 0.2),
+  ses_cat = sample(x = c("1_low", "2_middle", "3_high"), size = n, replace = TRUE, prob = c(0.2 , 0.4, 0.4)),
+  copd_cat = rbinom(n, size = 1, prob = 0.3 + 0.5*smoking_cat)
+  ) %>%  
+  # bring data in right format
+  dplyr::mutate(across(ends_with("num"), as.numeric)) %>% 
+  dplyr::mutate(across(ends_with("num"), function(x) round(x, digits = 2)))
+```
+
+In the first step, we create a dataset with `r formatC(nrow(sim_covar), format = "fg", big.mark = ",")` patients and `r ncol(sim_covar)` variables with the following distributions.
+
+```{r, distributions_covars}
+sim_covar %>% 
+  gtsummary::tbl_summary(by = "exposure") %>% 
+  gtsummary::add_difference()
+```
+
+The following table illustrates the odds of exposure assignment.
+
+```{r, pr(exposure)_tbl}
+exposure_form <- as.formula(paste("exposure ~ ", paste(colnames(sim_covar %>% dplyr::select(-exposure)), collapse = " + ")))
+
+exposure_fit <- stats::glm(
+  exposure_form,
+  data = sim_covar,
+  family = "binomial"
+  )
+
+exposure_fit %>% 
+  gtsummary::tbl_regression(exponentiate = T)
+```
+
+Fitting a generalized linear model and assessing the probability of treatment assignment, the above constellation of odds results in the following simulated distributions depicting propensities of treatment initiation (aka propensity scores).
+
+```{r, pr_treatment_assignment, fig.cap="Treatment assignment probabilities."}
+# compute propensity score
+exposure_plot <- sim_covar %>% 
+  dplyr::mutate(ps = fitted(exposure_fit))
+
+# plot density
+exposure_plot %>% 
+  ggplot2::ggplot(ggplot2::aes(x = ps, fill = factor(exposure))) +
+  ggplot2::geom_density(alpha = .5) +
+  ggplot2::theme_bw() +
+  ggplot2::labs(
+    x = "Pr(exposure)",
+    y = "Density",
+    fill = "Exposed"
+  )
+```
+
+### Simulate time-to-event
+
+Next, we simulate a time-to-event outcome for `overall survival`. For this, the `simsurv` package is used with the following assumptions:
+
+-   Parametric event times, following an exponential distribution
+-   Max event times: 5 years of follow-up
+-   Event times depend on some (time-fixed/baseline) covariate effects as listed in the table above
+-   beta coefficients for the outcome model as defined in the following
+
+```{r, betas_outcome_generation}
+betas_os <- c(
+  exposure = log(1),
+  age_num = log(1.05),
+  female_cat = log(0.94),
+  ecog_cat = log(1.25),
+  smoking_cat = log(1.3),
+  physical_cat = log(0.79),
+  egfr_cat = log(0.5),
+  alk_cat = log(0.91),
+  pdl1_num = log(0.98),
+  histology_cat = log(1.15)
+  )
+
+betas_os %>% 
+  as.data.frame() %>% 
+  dplyr::transmute(logHR = round(`.`, 2)) %>% 
+  tibble::rownames_to_column(var = "Covariate") %>% 
+  dplyr::mutate(HR = round(exp(logHR), 2)) %>% 
+  gt::gt()
+```
+
+```{r, outcome_generation, message=FALSE}
+set.seed(seed_value)
+
+sim_df <- sim_covar %>% dplyr::bind_cols(
+  simsurv::simsurv(
+    dist = "exponential",
+    lambdas = 0.05,
+    betas = betas_os,
+    x = sim_covar,
+    maxt = 5 
+    )
+  ) %>% 
+  dplyr::select(-id)
+```
+
+#### Kaplan-Meier estimates
+
+The simulation resulted in the following crude Kaplan-Meier estimates for 1-yr and 5-yr survival. Unadjusted estimates belown are shown overall and by exposure for overall survival.
+
+```{r, km_estimates}
+km_overall <- survival::survfit(survival::Surv(eventtime, status) ~ 1, data = sim_df)
+km_exposure <- survival::survfit(survival::Surv(eventtime, status) ~ exposure, data = sim_df)
+
+gtsummary::tbl_survfit(
+  list(km_overall, km_exposure),
+  times = c(1, 5),
+  label_header = "**{time} Years**"
+  )
+```
+
+Given, that the true exposure effect is null, the crude model is severely biased as we can see even more clearly in the crude Kaplan-Meier curve.
+
+```{r}
+km_exposure <- survival::survfit(survival::Surv(eventtime, status) ~ exposure, data = sim_df)
+
+survminer::ggsurvplot(
+  km_exposure, 
+  data = sim_df,
+  conf.int = TRUE,
+  surv.median.line = "hv",
+  palette = "jco",
+  xlab = "Time [Years]",
+  legend.labs = c("Comparator", "Exposure of interest")
+  )
+```
+
+#### Cox proportional hazards
+
+After adjusting, the simulated data results in the following hazard ratio (HR) estimates.
+
+```{r, Cox_estimates}
+cox_lhs <- "survival::Surv(eventtime, status)"
+cox_rhs <- paste(colnames(sim_covar), collapse = " + ")
+cox_form = as.formula(paste(cox_lhs, "~ exposure +", cox_rhs))
+
+cox_fit <- survival::coxph(cox_form, data = sim_df)
+
+cox_fit %>% 
+  gtsummary::tbl_regression(exponentiate = T)
+```
+
+## Introduce missingness
+
+```{r}
+smdi_data_complete <- sim_df
+```
+
+In many different quantitative disciplines from classic epidemiology to machine and deep learning there is an increasing interest in utilizing electronic health records (EHR) to develop prognostic/predictive models or study the comparative effectiveness and safety of medical interventions.
+Especially information on variables which are not readily available in other datasets (e.g.
+administrative claims) are of high interest, including vital signs, biomarkers and lab data.
+However, these covariates are often just partially observed for various reasons:
+
+-   Physician did simply not perform/order a certain test
+-   Certain measurements are just collected for particularly sick patients
+-   Information is 'hiding' in unstructured records, e.g. clinical notes
+
+To illustrate `smdi's` main functions using this dataset, we introduce some missingness to relevant covariates, which are critical confounders of the causal exposure-outcome relationship in the `smdi_data` dataset.
+
+In order to introduce missingness following different missingness mechanisms, we use the `ampute` function of the [mice](https://amices.org/mice/index.html) package.
+An excellent tutorial on this very flexible and elegant function can be found [here](https://rianneschouten.github.io/mice_ampute/vignette/ampute.html).
+
+In brief, we introduce missingness by...
+
+-   Determining a [missingess pattern](https://rianneschouten.github.io/mice_ampute/vignette/ampute.html#Patterns)
+-   Creating a [weight vector](https://rianneschouten.github.io/mice_ampute/vignette/ampute.html#Weights), using fitted linear models for the probability of observations becoming missing
+-   Specifying a type of logistic probability distribution for the missingness weights
+-   Defining an overall missingness proportion
+
+```{r}
+# prepare a placeholder df for missing simulation
+# we do not consider ses_cat
+tmp <- smdi_data_complete %>% 
+  dplyr::select(-c(ses_cat))
+
+# determine missingness pattern template
+miss_pattern <- rep(1, ncol(tmp))
+```
+
+## Missing complete at random
+
+`ecog_cat`[^1] will be set to missing according to the following specification:
+
+[^1]: The Eastern Cooperative Oncology Group (ECOG) performance score is a clinical measure for how the cancer disease affects the daily living abilities of the patient and is often used as a patient inclusion criterion for clinical trials.
+The scale ranges from 0 to 5 and typically only patients with 0 and 1 are eligible.
+Let's assume for our example, we already have a subset of such a pre-selected clinical trial-like cohort, but we still want to adjust for a baseline ECOG of 0 and 1 (definitions are taken from <https://ecog-acrin.org/resources/ecog-performance-status/>
+
+-   Missingness pattern: Only `ecog_cat` will be set to missing
+-   Weight vector: There are no weights, since the missingness is not dependent on any other observed or unobserved covariates
+-   Type: Not applicable
+
+```{r, mcar}
+# specify missingness pattern
+# (0 = set to missing, 1 = remains complete)
+mcar_col <- which(colnames(tmp)=="ecog_cat")
+miss_pattern_mcar <- replace(miss_pattern, mcar_col, 0)
+
+miss_prop_mcar <- .35
+
+set.seed(42)
+smdi_data_mcar <- mice::ampute(
+  data = tmp,
+  prop = miss_prop_mcar,
+  mech = "MCAR",
+  patterns = miss_pattern_mcar,
+  bycases = TRUE
+  )$amp %>% 
+  dplyr::select(ecog_cat)
+
+smdi_data_mcar %>% 
+  dplyr::select(ecog_cat) %>% 
+  dplyr::mutate(ecog_cat = forcats::fct_na_value_to_level(factor(ecog_cat), level = "missing")) %>% 
+  gtsummary::tbl_summary()
+```
+
+## Missing at random
+
+`egfr_cat` will be set to missing according to the following specification:
+
+-   Missingness pattern: Only `egfr_cat` will be set to missing
+-   Weight vector: all covariates (except `egfr_cat` itself and `ses_cat`) contribute with equal weights as linear predictors for the probability of observations becoming missing
+-   Type: higher weighted sum scores (i.e. higher odds of having an `egfr_cat` mutation) will have a larger probability of becoming incomplete
+-   Define an overall missingness proportion of \~40%
+
+```{r, mar}
+# specify missingness pattern
+# (0 = set to missing, 1 = remains complete)
+mar_col <- which(colnames(tmp)=="egfr_cat")
+miss_pattern_mar <- replace(miss_pattern, mar_col, 0)
+
+# weights to compute missingness probability 
+# by assigning a non-zero value
+miss_weights_mar <- rep(1, ncol(tmp))
+miss_weights_mar <- replace(miss_weights_mar, mar_col, 0)
+
+miss_prop_mar <- .4
+
+set.seed(42)
+smdi_data_mar <- mice::ampute(
+  data = tmp,
+  prop = miss_prop_mar,
+  mech = "MAR",
+  patterns = miss_pattern_mar,
+  weights = miss_weights_mar,
+  bycases = TRUE,
+  type = "RIGHT"
+  )$amp
+
+smdi_data_mar %>% 
+  dplyr::select(egfr_cat) %>% 
+  dplyr::mutate(egfr_cat = forcats::fct_na_value_to_level(factor(egfr_cat), level = "missing")) %>%
+  gtsummary::tbl_summary()
+```
+
+## Missing not at random - value
+
+-   Missingness pattern: Only `pdl1_num` will be set to missing
+-   Weight vector: Only `pdl1_num` itself (by a non-zero value) is the linear predictor for the probability of observations becoming missing
+-   Type: Specify a type of logistic probability distribution for the missingness weights, so that cases with low weighted sum scores (i.e.
+lower `pdl1_num` expression) will have a larger probability of becoming incomplete
+-   Define an overall missingness proportion of \~20%
+
+```{r, create_mnar_v}
+# determine missingness pattern
+mnar_v_col <- which(colnames(tmp)=="pdl1_num")
+miss_pattern <- rep(1, ncol(tmp))
+miss_pattern_mnar_v <- replace(miss_pattern, mnar_v_col, 0)
+
+# weights to compute missingness probability 
+# by assigning a non-zero value
+# MNAR_v: covariate itself is only predictor
+miss_weights_mnar_v <- rep(0, ncol(tmp))
+miss_weights_mnar_v <- replace(miss_weights_mnar_v, mnar_v_col, 1)
+
+miss_prop_mnar_v <- .2
+
+set.seed(42)
+smdi_data_mnar_v <- mice::ampute(
+  data = tmp,
+  prop = miss_prop_mnar_v,
+  mech = "MNAR",
+  patterns = miss_pattern_mnar_v,
+  weights = miss_weights_mnar_v,
+  bycases = TRUE,
+  type = "LEFT"
+  )$amp
+
+smdi_data_mnar_v %>% 
+  dplyr::select(pdl1_num) %>% 
+  gtsummary::tbl_summary()
+```
+
+## Assemble final dataset
+
+```{r}
+smdi_data <- smdi_data_complete %>% 
+  dplyr::select(-c(ecog_cat, egfr_cat, pdl1_num)) %>% 
+  dplyr::bind_cols(ecog_cat = smdi_data_mcar$ecog_cat, egfr_cat = smdi_data_mar$egfr_cat, pdl1_num = smdi_data_mnar_v$pdl1_num) %>% 
+  mutate(across(ends_with("cat"), as.factor))
+```
 
@@
 #| out-width: 115%
+#| eval: false
 

### 2023-07-22 manuscript/manuscript.qmd
updated renv lock
added figures and modified text

@@
 library(gtsummary)
+library(flextable)
+library(officer)
+library(captioner)
 library(smdi)
+
+table_cap <- captioner(prefix = "Table")
 ```
@@
 
-[**Keywords:**]{.underline} Missing data, Confounder, EHR, R, Software,
+[**Keywords:**]{.underline} Missing data, Confounder, EHR, R, Software, Real-World Evidence
 
@@
 
-[**Disclosures/COI:res/COI:**]{.underline} The FDA approved the study protocol, statistical analysis plan and reviewed and approved this manuscript.
+[**Disclosures/COI:**]{.underline} The FDA approved the study protocol, statistical analysis plan and reviewed and approved this manuscript.
 Coauthors from the FDA participated in the results interpretation and in the preparation and decision to submit the manuscript for publication.
@@
 In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019].
-Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
-These methods comprise standard procedures such as comparing distributions of baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
+Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms (see definitions box).
+These methods comprise standard procedures such as comparing distributions of baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
 However, these methods have so far been only described and tested in isolation from each other and no principled approach exists.
 In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed in both RWE and RCTs [@carroll2020; @wood2004; @harel2012].
+
+::: {#taxonomies .callout-note}
+## Definitions: Basic missing data taxonomies.
+
+### Patterns (taken from Van Buuren [@vanbuuren2018])
+
+-   Monotone pattern: If Y~j~ is the j^th^ column in a dataset *Y*, a missing data pattern is said to be *monotone* if the variables Y~j~ can be ordered such that if Y~j~ is missing then all variables Y~k~ with *k \> j* are also missing.
+This can occur, for example, in longitudinal studies with drop-out or with clinical labs which are typically measured together as part of a lab panel (e.g., renal or liver panel).
+
+-   Non-monotone pattern: If the pattern is not monotone, it is called *non-monotone* or *general*.
+
+### Mechanisms (taken from Sondhi and Weberpals et al. [@sondhi2023a])
+
+-   **MCAR**: The missingness does not depend on any other observed or unobserved covariate(s).
+
+-   **MCAR**: The missingness depends and can be explained by other observed covariates.
+
+-   **MNAR**: The missingness depends on unoberserved covariate(s).
+For example, the missingness may be explained by other covariate(s) which is/are not observed in the underlying dataset (MNAR~unmeasured~).
+The missingness can also just dependent on the actual value of the partially observed covariate itself (MNAR~value~).
+:::
 
@@
 
-## Problem formalization
+## Problem Formalization
 
@@
 
-## Theoretical background and main package functions
+## Theoretical Background and Main Package Functions
 
-For the implementation of these routine missing data diagnostics, we categorized the main functions of the `smdi` R package into three group diagnostics based on their general analytic prroperties (Table 1).
+For the implementation of these routine missing data diagnostics, we categorized the main functions of the `smdi` R package into three group diagnostics based on their general analytic properties (**`r table_cap("tbl1", display = "cite")`**).
 In this section, we will explain the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and the interpretation of results.
 Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates (one-row-per-patient format).
-Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the function.
+Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions.
 Throughout all functions, the user has the option to specify a vector with the column name(s) of the partially observed covariate(s) that should be investigated (`covar` parameter).
-If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.
+If nothing is specified, all functions consider automatically any variable in the dataframe that exhibits at least one missing value.
 
-### Group 1 diagnostics
+### Group 1 Diagnostics
 
@@
 
-To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics in the dataset [@schober2019; @austin2011; @tableone].
-The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs (which can be flexibly chosen using the `median` parameter) along with the corresponding minimum-maximum ASMD range.
+To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone].
+The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs (which can be flexibly chosen using the `median` parameter) along with corresponding minimum and maximum AMSDs observed.
 The *asmd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate.
@@
 
-### Group 2 diagnostics
+### Group 2 Diagnostics
 
@@
 It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model.
-For tuning the random forest model, the `smdi_rf()` function has a few parameters that can be specified by the user for optimization such as the number of trees to grow (default is 1,000 trees), the ratio of the split between train and test datasets (default is a 70/30 split) and the number of cores to parallelize the computation on [@parallel], since this function may be very time consuming, especially with larger datasets.
+For tuning the random forest model, the `smdi_rf()` function has a few parameters that can be specified by the user for optimization such as the number of trees to grow (`ntree`; default is 1,000 trees), the ratio of the split between train and test datasets (`train_test_ratio`; default is a 70/30 split) and the number of cores to parallelize the computation on (`n_cores`, default is 1), since this function can be very time consuming, especially with larger datasets.
 
-### Group 3 diagnostics
+### Group 3 Diagnostics
 
-Finally, the third group diagnostics with the `smdi_outcome()` function examines association of the missingness indicator of the partially observed covariate and the outcome under study.
+Finally, the third group diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the partially observed covariate and the outcome under study.
 The function will compute both a univariate model and a model adjusted for all other covariates included in the dataset.
@@
 Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes.
-Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be displayed exponentiated or not.
+Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be exponentiated or not.
 The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each partially observed covariate.
 
-## Exemplary RWE study and data generation
+## Exemplary RWE Study and Data Generation
 
-To showcase the practical implementation of the functions provided in `smdi`, the package comes with simulated example EHR dataset focusing on an exemplary RWE study in the field of oncology with `r formatC(nrow(smdi_data), format = "fg", big.mark = ",")` lung cancer patients (Table 2).
+### Study Population
 
-For this exemplary RWE study, we compared two systemic antineoplastic durg treatment regimens (`Drug A` versus `Drug B`).
-The true outcome of interest was overall survival (OS) which was defined as the time from treatment initiation to death due to any reason or censoring with a truncated follow-up time of 5 years (Supplementary Table and Supplementary Figure ).
-To simula The desired treatment effect estimate was defined with a hazard ratio (HR) of 1.0 and we assumed that the proportional hazards assumption holds.
+To showcase the practical implementation and suggested workflow when working with `smdi`, the package comes with a *de novo* simulated example EHR dataset that mimics a new-user cohort of `r formatC(nrow(smdi_data), format = "fg", big.mark = ",")` lung cancer patients initiating two comparable antineoplastic drug treatment regimens.
+The aim of the exemplary RWE study is to compare the comparative effectiveness of these two drug regimens in presence of confounding and three partially observed covariates, each following a different missingness mechanism.
+An overview of the dataset and baseline patient characteristics are provided in **`r table_cap("tbl2", display = "cite")`**.
 
-### Covariates
+### Simulation Design
+
+The data generation process is described in detail in the Supplementary Methods.
+In brief, for this exemplary RWE study, we simulated a binary exposure (*Drug A versus Drug B*), a time-to-event outcome (*overall survival,* defined *as* as the time from treatment initiation until all-cause mortality or censoring) and 11 covariates, seven of which were true confounders (i.e., associated with both exposure and outcome), two were prognostic factors (i.e., associated only with the outcome but not with exposure) and two covariates which were neither associated with the exposure nor the outcome (**`r table_cap("tbl3", display = "cite")`**).
+We generated the survival times for the outcome assuming parametric event times following an exponential distribution and a truncated follow-up time of 5 years (Supplementary Table and Supplementary Figure ) [@simsurv].
+The desired true treatment effect estimate was defined with a hazard ratio (HR) of 1.0.
+The strength of association between each covariate and the exposure and the outcome is displayed in Supplementary Table x and y, respectively.
+
+### Missingness generation
+
+To simulate missingness
 
@@
 
+`r table_cap(name = "tbl1", "Overview of the main functions in smdi to characterize potential underlying missingness mechanisms.")`
+
 ```{r}
-#| label: tbl_overview
-#| eval: true
 #| echo: false
@@
   gt() %>% 
-  tab_caption(caption = "tbl-overview") %>% 
-  tab_header(
-    title = "Overview of the main functions in smdi."
-    ) %>%
   tab_style(
@@
 
+`r table_cap(name = "tbl2", "Patient characteristics of the exemplary cohort used to showcase the smdi package.")`
+
 ```{r}
-#| label: tbl_cohort
-#| eval: true
 #| echo: false
@@
 
-tbl_cohort <- smdi_data %>% 
+# set compact theme 
+theme_gtsummary_compact(font_size = 11)
+
+tbl2 <- smdi_data %>% 
   select(-c(eventtime, status)) %>% 
   mutate(exposure = ifelse(exposure == 0, "Drug A", "Drug B")) %>% 
-  mutate(across(c(female_cat, smoking_cat, physical_cat, alk_cat, histology_cat, copd_cat), ~ifelse(.x == "0", 0, 1))) %>% 
+  mutate(
+    across(
+      c(
+        female_cat, 
+        smoking_cat, 
+        physical_cat, 
+        alk_cat, 
+        histology_cat, 
+        copd_cat
+        ), 
+      ~ifelse(.x == "0", 0, 1)
+      )
+    ) %>% 
   tbl_summary(
@@
   add_overall() %>% 
-  add_difference()
+  add_p() %>% 
+  as_flex_table() %>% 
+  font(fontname = "Calibri") %>% 
+  align(align = "left")
 
-tbl_cohort
+tbl2
 ```
-:::
+
+{{< pagebreak >}}
+
+`r table_cap(name = "tbl3", "Overview of covariates of de novo simulated cohort used to showcase the smdi package.")`
+
+```{r}
+#| echo: false
+
+read_excel(
+  path = here('tables', 'tables_manual.xlsx'),
+  sheet = "covariates"
+  ) %>% 
+  select(-`Variable Name`) %>% 
+  gt() %>% 
+  tab_style(
+    cell_text(weight = "bold"),
+    locations = cells_column_labels(columns = everything())
+    )
+```
 
@@
 ```{r}
-#| label: fig-RStudio-gui
-#| fig-cap: "Steps to clone a remote repository using the RStudio graphical user interface."
-#| eval: false
+#| label: fig-workflow
+#| fig-cap: "Overview of all `smdi` functions and suggested workflow to conduct missing data diagnostics.
+*gg_miss_upset() and md.pattern() are re-exports of the naniar and mice package, respectively."
 #| echo: false
-#| out-width: 50%
-knitr::include_graphics(here::here("figures", ""))
+#| out-width: 110%
+knitr::include_graphics(here::here("figures", "workflow.jpg"))
 ```
@@
 {{< pagebreak >}}
+
+```{r}
+#| label: fig-univariate
+#| fig-cap: "Complete case analysis."
+#| echo: false
+#| eval: false
+#| out-width: 110%
+
+library(survival)
+library(broom)
+suppressPackageStartupMessages(library(mice))
+library(survival)
+
+# specifiy form
+cox_lhs <- "survival::Surv(eventtime, status)"
+cox_rhs <- paste(colnames(smdi_data %>% select(-c(eventtime, status))), collapse = " + ")
+cox_form <- as.formula(paste(cox_lhs, "~", cox_rhs))
+
+# fit univariate
+cox_univariate <- survival::coxph(survival::Surv(eventtime, status) ~ exposure, 
+                                  data = smdi_data) %>% 
+  {. ->> cox_n} %>% 
+  broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>% 
+  filter(term == "exposure") %>% 
+  mutate(
+    method = "Univariate",
+    n = formatC(cox_n$n, big.mark = ",", format = "fg"),
+    events = formatC(cox_n$nevent, big.mark = ",", format = "fg"),
+    ) %>% 
+  select(method, n, events, estimate, conf.low, conf.high)
+
+# fit cox for complete case analysis
+cox_cca <- survival::coxph(cox_form, data = smdi_data) %>% 
+  {. ->> cox_n} %>% 
+  broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>% 
+  filter(term == "exposure") %>% 
+  mutate(
+    method = "Complete case analysis",
+    n = formatC(cox_n$n, big.mark = ",", format = "fg"),
+    events = formatC(cox_n$nevent, big.mark = ",", format = "fg")
+    ) %>% 
+  select(method, n, events, estimate, conf.low, conf.high)
+
+# fit multiple imputation model and cox model
+cox_mi <- mice::futuremice(
+  data = smdi_data,
+  method = "rf",
+  rfPackage = "randomForest",
+  m = 50,
+  n.core = parallel::detectCores(),
+  parallelseed = 42,
+  print = FALSE
+  ) %>% 
+  with(expr = survival::coxph(formula(paste(format(cox_form), collapse = "")))) %>%
+  mice::pool() %>%
+  {. ->> cox_n} %>% 
+  broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>% 
+  filter(term == "exposure") %>% 
+  mutate(
+    method = "Multiple imputation",
+    n = formatC(mean(cox_n$glanced$n), big.mark = ",", format = "fg"),
+    events = formatC(mean(cox_n$glanced$nevent), big.mark = ",", format = "fg")
+    ) %>% 
+  select(method, n, events, estimate, conf.low, conf.high)
+
+bind_rows(cox_univariate, cox_cca, cox_mi) %>% 
+  mutate(across(where(is.numeric), ~formatC(.x, digits = 2, format = "f"))) %>% 
+  gt()
+```
+
+{{< pagebreak >}}
+:::

### 2023-07-31 manuscript/manuscript.qmd
minor formatting changes

@@
 
-[**Word count:**]{.underline} xxxx words
+[**Word count:**]{.underline} xxxx words/2000 words
 
-[**Tables:**]{.underline}
+[**Tables:**]{.underline} **x/2**
 
-[**Figures:**]{.underline}
+[**Figures:**]{.underline} **x/3**
 
@@
 [**Analytical code and data sharing statement:**]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2.
-All R code and materials can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>.
-Detailed information on packages and versions can be found in the `renv.lock` file in the linked repository.
-The R package and data presented in this study can be downloaded from <https://janickweberpals.gitlab-pages.partners.org/smdi/>.
+All R code, materials and depedency information can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>.
+The R package presented in this study can be downloaded from CRAN via `install.packages("smdi")` or from <https://janickweberpals.gitlab-pages.partners.org/smdi/>.
 
@@
 # Abstract {.unnumbered}
+
+xxx words/150 words
 

### 2023-08-01 manuscript/manuscript.qmd
introduced headings according to general instructions page

@@
 library(captioner)
-library(smdi)
 
@@
 
-# Background
+# Background and Significance
 
@@
 
+# Objective
+
 To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we here present and demonstrate the use of the `smdi` (structural missing data investigations) R package [@smdi].
@@
 Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for specific missing data approaches hold [@pedersen2017; @lee2013].
+
+## **Package details and installation instructions**
+
+...
 

### 2023-08-02 manuscript/manuscript.qmd
added objectives

@@
 Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms (see definitions box).
-These methods comprise standard procedures such as comparing distributions of baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
+These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
 However, these methods have so far been only described and tested in isolation from each other and no principled approach exists.
-In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed in both RWE and RCTs [@carroll2020; @wood2004; @harel2012].
+In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed [@carroll2020; @wood2004; @harel2012].
 
@@
 To overcome these shortcomings, we [@SentinelCI3] have recently developed and evaluated a principled approach combining a range of missing data diagnostics in an US EHR-claims database linkage [@weberpals2023].
-The results of this large-scale study revealed that the combination of these diagnostics characterized simulated missing data mechanisms well and provided helpful insights for the appropriate choice of analytic methods to handle the partially observed confounder data (e.g., missing data imputation).
+The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle the partially observed confounder data (e.g., missing data imputation).
 
@@
 
-To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we here present and demonstrate the use of the `smdi` (structural missing data investigations) R package [@smdi].
-To that end, we first provide an overview on the package's main functions including the underlying theoretical assumptions.
-We then demonstrate the practical application of the package by illustrating and end-to-end workflow with a hypothetical RWE study using a simulated oncology EHR dataset that comes bundled with the package.
+To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
+The goal of `smdi` is to provide a toolbox of different diagnostics to scrutinize partially observed confounder data which had not been implemented elsewhere yet and complement other analytic R packages regularly used in RWE studies.
 
@@
 
-## Problem Formalization
-
-As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data (RWD) are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., EHR).
-Hence, confounders and prognostic factors, which need to be balanced between treatment groups in RWE studies, are usually not available for all patients at all necessary time points [@toh2012a].
-If the underlying mechanism for the missingness of such covariates is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to be missing, this can lead to bias in the resulting effect estimates for the studied treatments under common missing data approaches like complete case analysis or imputation [@groenwold2020].
-Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for specific missing data approaches hold [@pedersen2017; @lee2013].
-
 ## **Package details and installation instructions**
 
-...
+The smdi R package was written in R language (version 4.2.1)
 
@@
 The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each partially observed covariate.
-
-## Exemplary RWE Study and Data Generation
-
-### Study Population
-
-To showcase the practical implementation and suggested workflow when working with `smdi`, the package comes with a *de novo* simulated example EHR dataset that mimics a new-user cohort of `r formatC(nrow(smdi_data), format = "fg", big.mark = ",")` lung cancer patients initiating two comparable antineoplastic drug treatment regimens.
-The aim of the exemplary RWE study is to compare the comparative effectiveness of these two drug regimens in presence of confounding and three partially observed covariates, each following a different missingness mechanism.
-An overview of the dataset and baseline patient characteristics are provided in **`r table_cap("tbl2", display = "cite")`**.
-
-### Simulation Design
-
-The data generation process is described in detail in the Supplementary Methods.
-In brief, for this exemplary RWE study, we simulated a binary exposure (*Drug A versus Drug B*), a time-to-event outcome (*overall survival,* defined *as* as the time from treatment initiation until all-cause mortality or censoring) and 11 covariates, seven of which were true confounders (i.e., associated with both exposure and outcome), two were prognostic factors (i.e., associated only with the outcome but not with exposure) and two covariates which were neither associated with the exposure nor the outcome (**`r table_cap("tbl3", display = "cite")`**).
-We generated the survival times for the outcome assuming parametric event times following an exponential distribution and a truncated follow-up time of 5 years (Supplementary Table and Supplementary Figure ) [@simsurv].
-The desired true treatment effect estimate was defined with a hazard ratio (HR) of 1.0.
-The strength of association between each covariate and the exposure and the outcome is displayed in Supplementary Table x and y, respectively.
-
-### Missingness generation
-
-To simulate missingness
 

### 2023-08-02 manuscript/manuscript.qmd
adding section on missing data patterns

@@
 However, these methods have so far been only described and tested in isolation from each other and no principled approach exists.
-In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed [@carroll2020; @wood2004; @harel2012].
+In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed in studies [@carroll2020; @wood2004; @harel2012].
 
@@
 To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
-The goal of `smdi` is to provide a toolbox of different diagnostics to scrutinize partially observed confounder data which had not been implemented elsewhere yet and complement other analytic R packages regularly used in RWE studies.
-
-{{< pagebreak >}}
+The goal of `smdi` is to provide a toolbox of different diagnostic analyses to scrutinize partially observed confounder data which had not been implemented elsewhere yet and complement other analytic R packages regularly used in RWE studies.
 
@@
 
-The smdi R package was written in R language (version 4.2.1)
+The `smdi` R package was written in R language (version 4.2.1) and using the devtools [@devtools] and roxygen [@roxygen2] packages.
+The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi)>, Posit Public Package Manager (<https://packagemanager.posit.co/client/#/repos/2/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via the `install.packages("smdi")` command.
+To ensure the quality and robustness of the package and corresponding functions, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment which are scheduled to execute on a daily basis.
 
-## Theoretical Background and Main Package Functions
+# Results {#sec-results}
 
-For the implementation of these routine missing data diagnostics, we categorized the main functions of the `smdi` R package into three group diagnostics based on their general analytic properties (**`r table_cap("tbl1", display = "cite")`**).
-In this section, we will explain the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and the interpretation of results.
-Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates (one-row-per-patient format).
+## Main Package Functions
+
+@fig-workflow illustrates the recommended workflow and functions that can be used to systematically approach diagnostics on partially observed confounders.
+
+In general, the workflow comprises two broader diagnostic categories: descriptives and pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms.
+This section covers the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and the interpretation of results.
+Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates.
 Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions.
@@
 
-### Group 1 Diagnostics
+### Descriptives and Pattern Diagnostics
+
+As a first step to explore the missingness in new datasets, the `smdi` package provides a few basic functions to describe and summarize missingness across all covariates.
+The `smdi_summarize()` function takes the dataframe as input and returns the amount and proportion of missing observations for partially observed covariates which can also be stratified by one or more grouping variables (e.g., by an exposure or outcome variable).
+The `smdi_vis()` function returns a corresponding bar chart plot.
+
+[@ruddle2022]
+
+### Inferential Three Group Diagnostics
+
+The main functions used to investigate the potential underlying missing data mechanisms are categorized into three group diagnostics based on their general analytic properties (**`r table_cap("tbl1", display = "cite")`**).
+
+#### Group 1 Diagnostics
 
@@
 
-### Group 2 Diagnostics
+#### Group 2 Diagnostics
 
@@
 
-### Group 3 Diagnostics
+#### Group 3 Diagnostics
 
@@
 The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each partially observed covariate.
-
-{{< pagebreak >}}
-
-# Results {#sec-results}
 
@@
 
-`r table_cap(name = "tbl2", "Patient characteristics of the exemplary cohort used to showcase the smdi package.")`
-
-```{r}
-#| echo: false
-#| message: false
-
-labels <- readxl::read_excel(
-  path = here("tables", "tables_manual.xlsx"),
-  sheet = "covariates"
-  )
-
-labels_list <- as.list(labels$`Variable Label`)
-names(labels_list) <- labels$`Variable Name`
-
-# set compact theme 
-theme_gtsummary_compact(font_size = 11)
-
-tbl2 <- smdi_data %>% 
-  select(-c(eventtime, status)) %>% 
-  mutate(exposure = ifelse(exposure == 0, "Drug A", "Drug B")) %>% 
-  mutate(
-    across(
-      c(
-        female_cat, 
-        smoking_cat, 
-        physical_cat, 
-        alk_cat, 
-        histology_cat, 
-        copd_cat
-        ), 
-      ~ifelse(.x == "0", 0, 1)
-      )
-    ) %>% 
-  tbl_summary(
-    by = "exposure", 
-    label = labels_list, 
-    missing = "ifany"
-    ) %>% 
-  add_overall() %>% 
-  add_p() %>% 
-  as_flex_table() %>% 
-  font(fontname = "Calibri") %>% 
-  align(align = "left")
-
-tbl2
-```
-
-{{< pagebreak >}}
-
-`r table_cap(name = "tbl3", "Overview of covariates of de novo simulated cohort used to showcase the smdi package.")`
-
-```{r}
-#| echo: false
-
-read_excel(
-  path = here('tables', 'tables_manual.xlsx'),
-  sheet = "covariates"
-  ) %>% 
-  select(-`Variable Name`) %>% 
-  gt() %>% 
-  tab_style(
-    cell_text(weight = "bold"),
-    locations = cells_column_labels(columns = everything())
-    )
-```
-
-{{< pagebreak >}}
-
 # Figures {.unnumbered}
@@
 ```
-
-{{< pagebreak >}}
-
-```{r}
-#| label: fig-univariate
-#| fig-cap: "Complete case analysis."
-#| echo: false
-#| eval: false
-#| out-width: 110%
-
-library(survival)
-library(broom)
-suppressPackageStartupMessages(library(mice))
-library(survival)
-
-# specifiy form
-cox_lhs <- "survival::Surv(eventtime, status)"
-cox_rhs <- paste(colnames(smdi_data %>% select(-c(eventtime, status))), collapse = " + ")
-cox_form <- as.formula(paste(cox_lhs, "~", cox_rhs))
-
-# fit univariate
-cox_univariate <- survival::coxph(survival::Surv(eventtime, status) ~ exposure, 
-                                  data = smdi_data) %>% 
-  {. ->> cox_n} %>% 
-  broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>% 
-  filter(term == "exposure") %>% 
-  mutate(
-    method = "Univariate",
-    n = formatC(cox_n$n, big.mark = ",", format = "fg"),
-    events = formatC(cox_n$nevent, big.mark = ",", format = "fg"),
-    ) %>% 
-  select(method, n, events, estimate, conf.low, conf.high)
-
-# fit cox for complete case analysis
-cox_cca <- survival::coxph(cox_form, data = smdi_data) %>% 
-  {. ->> cox_n} %>% 
-  broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>% 
-  filter(term == "exposure") %>% 
-  mutate(
-    method = "Complete case analysis",
-    n = formatC(cox_n$n, big.mark = ",", format = "fg"),
-    events = formatC(cox_n$nevent, big.mark = ",", format = "fg")
-    ) %>% 
-  select(method, n, events, estimate, conf.low, conf.high)
-
-# fit multiple imputation model and cox model
-cox_mi <- mice::futuremice(
-  data = smdi_data,
-  method = "rf",
-  rfPackage = "randomForest",
-  m = 50,
-  n.core = parallel::detectCores(),
-  parallelseed = 42,
-  print = FALSE
-  ) %>% 
-  with(expr = survival::coxph(formula(paste(format(cox_form), collapse = "")))) %>%
-  mice::pool() %>%
-  {. ->> cox_n} %>% 
-  broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>% 
-  filter(term == "exposure") %>% 
-  mutate(
-    method = "Multiple imputation",
-    n = formatC(mean(cox_n$glanced$n), big.mark = ",", format = "fg"),
-    events = formatC(mean(cox_n$glanced$nevent), big.mark = ",", format = "fg")
-    ) %>% 
-  select(method, n, events, estimate, conf.low, conf.high)
-
-bind_rows(cox_univariate, cox_cca, cox_mi) %>% 
-  mutate(across(where(is.numeric), ~formatC(.x, digits = 2, format = "f"))) %>% 
-  gt()
-```
-
-{{< pagebreak >}}
 :::

### 2023-08-03 manuscript/manuscript.qmd
stopped at paragraph about set visualization
finalized main body text draft; discussion next

@@
 library(captioner)
+library(smdi)
+suppressPackageStartupMessages(library(ggimage))
 
@@
 
-::: {.callout-note icon="false"}
-## Highlights
+::: {.callout-tip icon="false"}
+## Lay Summary
 
@@
 However, these methods have so far been only described and tested in isolation from each other and no principled approach exists.
-In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed in studies [@carroll2020; @wood2004; @harel2012].
+In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and not often performed [@carroll2020; @wood2004; @harel2012].
+
+{{< pagebreak >}}
 
@@
 To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
-The goal of `smdi` is to provide a toolbox of different diagnostic analyses to scrutinize partially observed confounder data which had not been implemented elsewhere yet and complement other analytic R packages regularly used in RWE studies.
+The goal of `smdi` is to provide a toolbox of different diagnostics to analyze the potentially underlying missingness patterns and mechanisms of partially observed confounders in RWE studies.
 
-# Methods
+# Materials and Methods
 
-## **Package details and installation instructions**
-
-The `smdi` R package was written in R language (version 4.2.1) and using the devtools [@devtools] and roxygen [@roxygen2] packages.
+The `smdi` R package was written in R language (version 4.2.1) employing the devtools [@devtools] and roxygen [@roxygen2] packages.
 The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi)>, Posit Public Package Manager (<https://packagemanager.posit.co/client/#/repos/2/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via the `install.packages("smdi")` command.
 To ensure the quality and robustness of the package and corresponding functions, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment which are scheduled to execute on a daily basis.
+Additional resources such as documentation, articles and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.
 
@@
 
-@fig-workflow illustrates the recommended workflow and functions that can be used to systematically approach diagnostics on partially observed confounders.
+@fig-workflow illustrates the recommended workflow and functions to systematically approach diagnostics on partially observed confounders.
 
-In general, the workflow comprises two broader diagnostic categories: descriptives and pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms.
-This section covers the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and the interpretation of results.
-Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates.
+In general, the workflow comprises two broader diagnostic categories: descriptives/pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms.
+In this section, we cover the principles behind the main package functions, the parameters a user can specify, the returned results and the interpretation of results.
+Examples are illustrated using a simulated EHR dataset that is part of the package (more details can be found under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
+Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates.
 Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions.
 Throughout all functions, the user has the option to specify a vector with the column name(s) of the partially observed covariate(s) that should be investigated (`covar` parameter).
-If nothing is specified, all functions consider automatically any variable in the dataframe that exhibits at least one missing value.
+If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.
 
@@
 As a first step to explore the missingness in new datasets, the `smdi` package provides a few basic functions to describe and summarize missingness across all covariates.
-The `smdi_summarize()` function takes the dataframe as input and returns the amount and proportion of missing observations for partially observed covariates which can also be stratified by one or more grouping variables (e.g., by an exposure or outcome variable).
-The `smdi_vis()` function returns a corresponding bar chart plot.
+The `smdi_summarize()` function takes the dataframe as input and returns the amount and proportion of missing observations, which can also be stratified by a grouping variable (e.g., by an exposure or outcome variable).
+The `smdi_vis()` function returns a corresponding bar chart plot (example @fig-examples a).
 
-[@ruddle2022]
+To visually inspect potential missing data patterns, we re-exported the `gg_miss_upset()` function of the naniar package [@naniar-2].
+This function uses a set visualization technique to visually infer potential (non-)monotone patterns based on the number of intersecting missing observations across all partially observed covariates [@ruddle2022].
+For example, a monotone pattern would be visually evident if, for a set of two or more variables (e.g., lab1, lab2), all or the majority of missing records would be observed for both labs simultaneously (example @fig-examples b).
 
@@
 
-The main functions used to investigate the potential underlying missing data mechanisms are categorized into three group diagnostics based on their general analytic properties (**`r table_cap("tbl1", display = "cite")`**).
+The core functions to infer potentially underlying missingness mechanisms are categorized into three group diagnostics based on their general analytic properties (**`r table_cap("tbl1", display = "cite")`**).
 
@@
 The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the partially observed covariate.
-According to Rubin's framework on inference and missing data [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates.
+According to Rubin's framework [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates.
 Consequently, significant differences in patient characteristics would be expected under a MAR mechanism between strata of patients with and without the partially observed covariate.
@@
 To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone].
-The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs (which can be flexibly chosen using the `median` parameter) along with corresponding minimum and maximum AMSDs observed.
+The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs (which can be flexibly chosen using the `median` parameter) along with a corresponding range of minimum and maximum AMSDs observed.
 The *asmd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate.
-For a graphical visualization of this, the function also creates a *ggplot2* graph [@ggplot2] illustrating the ASMDs for each compared patient characteristic in descending order.
+For a graphical visualization of this, the function returns a graph illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2] .
 
 The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test.
-Hotelling's test [@hotelling1931; @Hotelling] formalizes this is a multivariate t-test for each partially observed covariate, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each partially observed covariate individually.
-In contrast, Little's test [@little1988; @naniar] (`smdi_little()`) computes a single global chi-square test statistic and p-value across all partially observed covariates with the null hypothesis that the data is MCAR.
+Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test for each partially observed covariate, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each partially observed covariate separately.
+In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all partially observed covariates with the null hypothesis that the data is MCAR.
 
@@
 
-The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` functions.
+The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
 This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors.
 If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism.
-In case of values close 0.5, this would rather indicate a random prediction and translate to potential MCAR or MNAR mechanisms.
+In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism.
 
-The function returns an object of class `rf` which generically displays an overview of the AUC values of all partially observed covariates.
-The AUC values are based on the prediction made in the respective test datasets.
-For more details, the `rf` object additionally includes a *ggplot2* graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy.
+The function returns an object of class *rf* which generically prints an overview of the AUC values of all partially observed covariates.
+The AUC value is based on the prediction made in the respective test dataset.
+The *rf* object further returns a graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d).
 This metric can be valuable for interpreting and identifying strong predictors of missingness.
 It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model.
-For tuning the random forest model, the `smdi_rf()` function has a few parameters that can be specified by the user for optimization such as the number of trees to grow (`ntree`; default is 1,000 trees), the ratio of the split between train and test datasets (`train_test_ratio`; default is a 70/30 split) and the number of cores to parallelize the computation on (`n_cores`, default is 1), since this function can be very time consuming, especially with larger datasets.
+In case of inflated AUC values (\>0.9), the function prompts a message to the user detailing the most important predictor.
+If another partially observed covariate is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to exclude this covariate and run the diagnostics for each partially observed covariate independently.
 
+```{=html}
+<!--
+
+To tune the random forest model hyperparameters, the `smdi_rf()` function has a few arguments that can be specified by the user for optimization such as the number of trees to grow (`ntree`; default is 1,000 trees), the ratio of the split between train and test datasets (`train_test_ratio`; default is a 70/30 split) and the number of cores to parallelize the computation on (`n_cores`, default is 1), since this function can be very time consuming, especially with larger datasets.
+
+-->
+```
 #### Group 3 Diagnostics
@@
 
+#### `smdi_diagnose()` to compute all three group diagnostics
+
+The `smdi_diagnose()` function enables a user to compute all of the three group diagnostics introduced above with just one function call.
+
+```{r}
+#| eval: false
+#| echo: true
+smdi_diagnose(
+  data = smdi_data,
+  covar = NULL, # NULL includes all covariates with at least one NA
+  model = "cox",
+  form_lhs = "Surv(eventtime, status)"
+  ) %>% 
+  smdi_style_gt()
+```
+
+The function will return an object of class *smdi* containing a table with the results of all diagnostics for each specified partially observed covariate and the global Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**).
+The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.
+
 {{< pagebreak >}}
@@
 # Discussion
+
+Missing data are ubiquitous in real-world databases and may lead to bias and spurious results if not handled appropriately.
+
+By comparing the resulting diagnostic parameters to expected estimates, based on previous simulation results [@weberpals2023], the diagnostics can provide valuable insights into underlying missingness structures and help elucidate if analytic approaches such as missing imputation analyses are viable options @fig-results.
 
@@
 
+`r table_cap(name = "tbl2", "Example output of the smdi_diagnose() function applied to the examplary smdi_data dataset.")`
+
+```{r}
+#| echo: false
+smdi_diagnose(
+  data = smdi_data,
+  covar = NULL, # NULL includes all covariates with at least one NA
+  model = "cox",
+  form_lhs = "Surv(eventtime, status)",
+  n_cores = 3,
+  ) %>% 
+  smdi_style_gt()
+```
+
+{{< pagebreak >}}
+
 # Figures {.unnumbered}
@@
 #| out-width: 110%
-knitr::include_graphics(here::here("figures", "workflow.jpg"))
+knitr::include_graphics(here::here("figures", "Figure_1_workflow.jpg"))
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-examples
+#| fig-cap: "Exemplary visual outputs of the a) smdi_vis(), b) gg_miss_upset(), c) smdi_asmd() and d) smdi_rf() functions, respectively."
+#| echo: false
+#| warning: false
+#| fig-dpi: 400
+#| fig-width: 15
+#| fig-height: 10
+#| out-width: 110%
+
+# a) smdi_vis()
+panel_a <- smdi_vis(data = smdi_data, strata = "exposure")
+panel_a <- panel_a + ggplot2::theme_bw(base_size = 15)
+
+# b) gg_miss_upset()
+# for panel B, we simulatea monotone missingness pattern
+# following an MCAR mechanism
+set.seed(42)
+
+data_monotone <- smdi_data_complete %>% 
+  mutate(
+    lab1 = rnorm(nrow(smdi_data_complete), mean = 5, sd = 0.5),
+    lab2 = rnorm(nrow(smdi_data_complete), mean = 10, sd = 2.25)
+    )
+
+data_monotone[3:503, "lab1"] <- NA
+data_monotone[1:500, "lab2"] <- NA
+
+panel_b <- as.ggplot(smdi::gg_miss_upset(data = data_monotone))
+
+#smdi_asmd
+asmd <- smdi_asmd(data = smdi_data)
+panel_c <- asmd$pdl1_num$asmd_plot
+
+#smdi_asmd
+rf <- smdi_rf(data = smdi_data, n_cores = 3)
+panel_d <- rf$ecog_cat$rf_plot
+
+panel <- cowplot::plot_grid(
+  panel_a, panel_b, panel_c, panel_d,
+  ncol = 2,
+  nrow = 2,
+  labels = c("a)", # % missing
+             "b)", # set visualization
+             "c)", # asmd
+             "d)" # rf
+             ),
+  label_size = 18,
+  label_colour = "black",
+  label_fontface = "bold",
+  vjust = 1.25,
+  hjust = -.25
+  )
+
+panel
+
+jpeg(here::here("figures", "Figure_2_visuals.jpg"), width = 16, height = 10, units = "in", res=400)
+panel
+invisible(dev.off())
 ```
 :::
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-results
+#| eval: true
+#| fig-cap: "Example of how smdi diagnostics can be applied to compute and compare diagnostic parameters of partially observed covariates to expected parameters of common missingness mechanisms based on a former large-scale simulation study [@weberpals2023]."
+#| echo: false
+#| out-width: 110%
+knitr::include_graphics(here::here("figures", "Figure_3_results.jpg"))
+```

### 2023-08-04 manuscript/manuscript.qmd
first draft of main body text before shortening

@@
 library(gt)
-library(gtsummary)
-library(flextable)
-library(officer)
 library(captioner)
 library(smdi)
-suppressPackageStartupMessages(library(ggimage))
+library(ggimage)
 
@@
 Email: [jweberpals\@bwh.harvard.edu](mailto:jweberpals@bwh.harvard.edu)
+
+[**Article type:**]{.underline} Application Note
 
@@
 
-In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019].
-Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms (see definitions box).
-These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
+In order to make an informed decision about the most appropriate analytic approach, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019].
+Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
+These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
 However, these methods have so far been only described and tested in isolation from each other and no principled approach exists.
-In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and not often performed [@carroll2020; @wood2004; @harel2012].
+In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently not often performed [@carroll2020; @wood2004; @harel2012].
 
@@
 
-### Patterns (taken from Van Buuren [@vanbuuren2018])
+### Patterns (adapted from Van Buuren [@vanbuuren2018])
 
@@
 
-### Mechanisms (taken from Sondhi and Weberpals et al. [@sondhi2023a])
+### Mechanisms [@sondhi2023a]
 
@@
 
--   **MCAR**: The missingness depends and can be explained by other observed covariates.
+-   **MAR**: The missingness depends and can be explained by other observed covariates.
 
@@
 
-To overcome these shortcomings, we [@SentinelCI3] have recently developed and evaluated a principled approach combining a range of missing data diagnostics in an US EHR-claims database linkage [@weberpals2023].
-The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle the partially observed confounder data (e.g., missing data imputation).
+In light of these limitations, we have recently developed and evaluated a principled approach with multiple missing data diagnostics in an US EHR-claims database linkage [@weberpals2023].
+The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data (e.g., missing data imputation).
 
@@
 To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
-The goal of `smdi` is to provide a toolbox of different diagnostics to analyze the potentially underlying missingness patterns and mechanisms of partially observed confounders in RWE studies.
+The goal of `smdi` was to provide a toolbox of different diagnostics to analyze the potentially underlying missingness patterns and mechanisms of POCs in RWE studies.
 
@@
 
-The `smdi` R package was written in R language (version 4.2.1) employing the devtools [@devtools] and roxygen [@roxygen2] packages.
-The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi)>, Posit Public Package Manager (<https://packagemanager.posit.co/client/#/repos/2/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via the `install.packages("smdi")` command.
+The `smdi` R package was written in R language (version 4.2.1).
+The package is available on CRAN ([https://cran.r-project.org/web/packages/smdi](https://cran.r-project.org/web/packages/smdi)){.uri}) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via the `install.packages("smdi")` command.
 To ensure the quality and robustness of the package and corresponding functions, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment which are scheduled to execute on a daily basis.
-Additional resources such as documentation, articles and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.
+Additional resources such as documentation and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.
 
@@
 
-@fig-workflow illustrates the recommended workflow and functions to systematically approach diagnostics on partially observed confounders.
+@fig-workflow illustrates the recommended workflow and functions to systematically approach diagnostics on POCs.
 
-In general, the workflow comprises two broader diagnostic categories: descriptives/pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms.
-In this section, we cover the principles behind the main package functions, the parameters a user can specify, the returned results and the interpretation of results.
-Examples are illustrated using a simulated EHR dataset that is part of the package (more details can be found under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
-Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates.
+The workflow is generally categorized into descriptives/pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms.
+In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and the results interpretation.
+Examples are illustrated using a simulated EHR dataset that is part of the package (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
+For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the POCs.
 Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions.
-Throughout all functions, the user has the option to specify a vector with the column name(s) of the partially observed covariate(s) that should be investigated (`covar` parameter).
+Throughout all functions, the user has the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter).
 If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.
@@
 To visually inspect potential missing data patterns, we re-exported the `gg_miss_upset()` function of the naniar package [@naniar-2].
-This function uses a set visualization technique to visually infer potential (non-)monotone patterns based on the number of intersecting missing observations across all partially observed covariates [@ruddle2022].
+This function uses a set visualization technique to visually infer potential (non-)monotone patterns based on the number of intersecting missing observations across all POCs [@ruddle2022].
 For example, a monotone pattern would be visually evident if, for a set of two or more variables (e.g., lab1, lab2), all or the majority of missing records would be observed for both labs simultaneously (example @fig-examples b).
@@
 
-The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the partially observed covariate.
+The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the POC.
 According to Rubin's framework [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates.
-Consequently, significant differences in patient characteristics would be expected under a MAR mechanism between strata of patients with and without the partially observed covariate.
+Consequently, significant differences in patient characteristics would be expected under a MAR mechanism between strata of patients with and without the POC.
 If the missingness depends only on unobserved factors (missing not at random \[MNAR\]) or does not depend on either observed or unobserved covariates (missing completely at random \[MCAR\]), differences should not be observable.
@@
 To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone].
-The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs (which can be flexibly chosen using the `median` parameter) along with a corresponding range of minimum and maximum AMSDs observed.
-The *asmd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate.
-For a graphical visualization of this, the function returns a graph illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2] .
+The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs along with a corresponding range of minimum and maximum AMSDs observed.
+The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC.
+For a graphical visualization of this, the function returns a figure illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2].
 
 The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test.
-Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test for each partially observed covariate, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each partially observed covariate separately.
-In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all partially observed covariates with the null hypothesis that the data is MCAR.
+Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test for each POC, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each POC separately.
+In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is MCAR.
 
@@
 The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
-This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors.
+This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the POC given observed covariates as the predictors.
 If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism.
@@
 
-The function returns an object of class *rf* which generically prints an overview of the AUC values of all partially observed covariates.
-The AUC value is based on the prediction made in the respective test dataset.
-The *rf* object further returns a graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d).
+The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC.
+The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio can be chosen by the user (`train_test_ratio` parameter).
+The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d).
 This metric can be valuable for interpreting and identifying strong predictors of missingness.
@@
 In case of inflated AUC values (\>0.9), the function prompts a message to the user detailing the most important predictor.
-If another partially observed covariate is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to exclude this covariate and run the diagnostics for each partially observed covariate independently.
+If another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to exclude this POC and run the diagnostics for each POC independently.
 
@@
 
-Finally, the third group diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the partially observed covariate and the outcome under study.
+The third group diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the POC and the outcome under study.
 The function will compute both a univariate model and a model adjusted for all other covariates included in the dataset.
-In previous simulations, we observed characteristic patterns for univariate and adjusted associations of the missing indicator and the outcome that matched simulated underlying missingness mechanisms [@weberpals2023].
-As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the partially observed covariate.
+In previous simulations, we observed characteristic patterns for univariate and adjusted associations of the missing indicator and the outcome that matched simulated missingness mechanisms [@weberpals2023].
+As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC.
 Under MAR, given that missingness can be explained by observed covariates, a potential spurious association in the univariate model disappeared after adjustment.
@@
 Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be exponentiated or not.
-The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each partially observed covariate.
+The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.
 
@@
 
-The `smdi_diagnose()` function enables a user to compute all of the three group diagnostics introduced above with just one function call.
+Finally, the `smdi_diagnose()` function enables a user to compute all three group diagnostics with just one function call.
 
@@
 #| echo: true
+
+# minimal example of a smdi_diagnose() function call
 smdi_diagnose(
@@
   model = "cox",
-  form_lhs = "Surv(eventtime, status)"
+  form_lhs = "Surv(eventtime, status)",
+  n_cores = 3 # number of cores to parallelize computations on
   ) %>% 
@@
 
-The function will return an object of class *smdi* containing a table with the results of all diagnostics for each specified partially observed covariate and the global Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**).
+The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and the global Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**).
 The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.
@@
 
-Missing data are ubiquitous in real-world databases and may lead to bias and spurious results if not handled appropriately.
+Missing data are ubiquitous in real-world databases and may lead to bias if not handled appropriately.
+We developed `smdi` to streamline routine diagnostics of missing data.
+By cross-checking the resulting diagnostic parameters to expected estimates, based on the outlined theoretical assumptions and previous simulation results [@weberpals2023], the diagnostics can provide valuable insights into underlying missingness structures and help elucidate if analytic approaches such as missing imputation analyses are viable options (example shown in @fig-results).
 
-By comparing the resulting diagnostic parameters to expected estimates, based on previous simulation results [@weberpals2023], the diagnostics can provide valuable insights into underlying missingness structures and help elucidate if analytic approaches such as missing imputation analyses are viable options @fig-results.
+The package also comes with limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data.
+Hence, it's important that diagnostic results are complemented with substantive expert knowledge to factor in how covariates are measured in routine care processes and contextualize potential reasons for missingness.
+
+## Conclusions
+
+The smdi package is a powerful and convenient tool to implement and carry out principled missing data diagnostics in RWE studies and improve the robustness of studies involving POCs by helping elucidate if certain analytic approaches are viable options for a given dataset.
 
@@
 #| fig-height: 10
-#| out-width: 110%
 
@@
 #| echo: false
-#| out-width: 110%
 knitr::include_graphics(here::here("figures", "Figure_3_results.jpg"))

### 2023-08-04 manuscript/manuscript.qmd
more main body text shaping and drafted abstracted; lay summary missing

@@
 
-[**Word count:**]{.underline} xxxx words/2000 words
+[**Word count:**]{.underline} 2,095 words/2,000 words
 
-[**Tables:**]{.underline} **x/2**
+[**Tables:**]{.underline} 2**/2**
 
-[**Figures:**]{.underline} **x/3**
+[**Figures:**]{.underline} 3**/3**
 
@@
 
-xxx words/150 words
+148 words/150 words
+
+**Objectives**
+
+Partially observed confounder data are a major challenge for the statistical analysis of electronic health records (EHR).
+While analytic approaches exist (e.g., multiple imputation), assumption on underlying missingness patterns and mechanisms have to hold.
+We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.
+
+**Materials and Methods**
+
+We developed the smdi (structural missing data diagnostics) R package based on structural assumptions of common missing data mechanisms in real-world data and results of a previous simulation study using a US EHR-claims database linkage.
+
+**Results**
+
+smdi enables users to run principled missing data investigations on partially observed confounders including descriptive analyses and diagnostics on potentially underlying missingness patterns and mechanisms.
+
+**Conclusions**
+
+smdi can give valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.
+smdi is freely available seamlessly integrates to the ecosystem of R packages for healthcare database analytics.
 
@@
 To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
-The goal of `smdi` was to provide a toolbox of different diagnostics to analyze the potentially underlying missingness patterns and mechanisms of POCs in RWE studies.
 
@@
 The `smdi` R package was written in R language (version 4.2.1).
-The package is available on CRAN ([https://cran.r-project.org/web/packages/smdi](https://cran.r-project.org/web/packages/smdi)){.uri}) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via the `install.packages("smdi")` command.
-To ensure the quality and robustness of the package and corresponding functions, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment which are scheduled to execute on a daily basis.
+The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`.
+To ensure the quality and robustness of the package, we implemented comprehensive unit tests with a coverage of 95.81% and established automated daily R CMD checks [@wickham2023r] via continuous integration and deployment.
 Additional resources such as documentation and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.
@@
 
-@fig-workflow illustrates the recommended workflow and functions to systematically approach diagnostics on POCs.
+@fig-workflow illustrates the recommended workflow to systematically approach diagnostics on POCs.
 
 The workflow is generally categorized into descriptives/pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms.
-In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and the results interpretation.
+In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and how these can be interpreted.
 Examples are illustrated using a simulated EHR dataset that is part of the package (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
-For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the POCs.
+For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent the variables relevant for the study, i.e., exposure, outcome, fully observed covariates and the POCs.
 Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions.
@@
 
-To visually inspect potential missing data patterns, we re-exported the `gg_miss_upset()` function of the naniar package [@naniar-2].
+To visually inspect potential missing data patterns, we re-exported the `gg_miss_upset()` function of the naniar package [@naniar].
 This function uses a set visualization technique to visually infer potential (non-)monotone patterns based on the number of intersecting missing observations across all POCs [@ruddle2022].
-For example, a monotone pattern would be visually evident if, for a set of two or more variables (e.g., lab1, lab2), all or the majority of missing records would be observed for both labs simultaneously (example @fig-examples b).
+For example, a monotone pattern would be visually evident if, for a set of two or more variables (e.g., lab1, lab2), all or the majority of missing records would be observed for both variables simultaneously (example @fig-examples b).
 
@@
 
-To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone].
+To quantify such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone].
 The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs along with a corresponding range of minimum and maximum AMSDs observed.
 The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC.
-For a graphical visualization of this, the function returns a figure illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2].
+For a graphical visualization of this, the function returns a plot illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2].
 
 The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test.
-Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test for each POC, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each POC separately.
+Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test for each POC, which means that `smdi_hotelling()` returns a test statistic and p-value for each each POC.
 In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is MCAR.
@@
 This metric can be valuable for interpreting and identifying strong predictors of missingness.
-It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model.
-In case of inflated AUC values (\>0.9), the function prompts a message to the user detailing the most important predictor.
+It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the model.
+In case of inflated AUC values (\>0.9), the function prompts a message to the user reporting the most important predictor.
 If another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to exclude this POC and run the diagnostics for each POC independently.
@@
 The function will compute both a univariate model and a model adjusted for all other covariates included in the dataset.
-In previous simulations, we observed characteristic patterns for univariate and adjusted associations of the missing indicator and the outcome that matched simulated missingness mechanisms [@weberpals2023].
+In preceding simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms [@weberpals2023].
 As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC.
-Under MAR, given that missingness can be explained by observed covariates, a potential spurious association in the univariate model disappeared after adjustment.
-If the missingness followed any MNAR mechanism, an association was observed in both univariate and adjusted models.
+Under MAR, given that missingness can be explained by observed covariates, a spurious association in the univariate model disappeared after adjustment.
+If the missingness followed any MNAR mechanism, an association was observed regardless of adjustment.
 
 Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes.
-Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be exponentiated or not.
+Besides the the regression type, a user needs to specify the column name that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be exponentiated or not.
 The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.
@@
 We developed `smdi` to streamline routine diagnostics of missing data.
-By cross-checking the resulting diagnostic parameters to expected estimates, based on the outlined theoretical assumptions and previous simulation results [@weberpals2023], the diagnostics can provide valuable insights into underlying missingness structures and help elucidate if analytic approaches such as missing imputation analyses are viable options (example shown in @fig-results).
+By cross-checking the resulting diagnostic parameters to expected estimates ( @fig-results, \[@weberpals2023\]), the diagnostics can provide valuable insights into underlying missingness patterns and mechanisms and help elucidate if analytic approaches such as imputation analyses are viable options.
 
@@
 The smdi package is a powerful and convenient tool to implement and carry out principled missing data diagnostics in RWE studies and improve the robustness of studies involving POCs by helping elucidate if certain analytic approaches are viable options for a given dataset.
+Through its design it seamlessly integrates with other R packages commonly used for clinical reporting [@gtsummary; @gt; @mice; @naniar; @ggplot2].
 
@@
 ```
-:::
 
@@
 ```
+:::

### 2023-08-07 README.Rmd
added readme

---
title: "README"
output: html_document
date: "2023-07-24"
---

## Background

This manuscript provides context and details about the smdi R package that was developed as a result of the findings of a comprehensive Sentinel Innovation Center project on `principled approaches on partially observed confounder data` in electronic health records.

More information about the package can be found under:
<https://janickweberpals.gitlab-pages.partners.org/smdi>

## Structure

-   .Rprofile - defines paths and activates `renv`, options for Posit R package manager
-   tables - main and supplementary tables (R objects and .docx format)
-   figures - main and supplementary figures (R objects and .docx format)
-   renv/renv.lock - `renv` directories to manage R dependencies and versions used in this simulation

## Dependencies

R package dependencies are managed through the `renv` package. All packages and their versions can be viewed in the lockfile `renv.lock`. All required packages and the appropriate versions can be installed by running the following command:

```{r, eval=FALSE}
renv::restore()
```

## Directory overview

```{r}
fs::dir_tree(recurse = 1)
```

```{r, include=FALSE, eval=FALSE}
rmarkdown::render(input = "README.Rmd", output_format = "md_document")
```

### 2023-08-07 manuscript/manuscript.qmd
added lay summary, changed to jamia citation style, minor modifications to manuscript (mostly shortening and wording)

@@
 ---
-title: "smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world databases"
+title: "smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies"
 
@@
 bibliography: references.bib
-csl: computer-methods-and-programs-in-biomedicine
+csl: jamia.csl
 
@@
 
-[**Word count:**]{.underline} 2,095 words/2,000 words
+[**Word count:**]{.underline} 2,070 words/2,000 words
 
-[**Tables:**]{.underline} 2**/2**
+[**Tables:**]{.underline} 2/2
 
-[**Figures:**]{.underline} 3**/3**
+[**Figures:**]{.underline} 3/3
 
-[**Supplementary material:**]{.underline}
+[**Supplementary material:**]{.underline} None
 
@@
 [**Analytical code and data sharing statement:**]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2.
-All R code, materials and depedency information can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>.
-The R package presented in this study can be downloaded from CRAN via `install.packages("smdi")` or from <https://janickweberpals.gitlab-pages.partners.org/smdi/>.
+All R code, materials and depedencies can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>.
+The R package presented in this study can be downloaded from CRAN via `install.packages("smdi")` or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.
 
@@
 
-148 words/150 words
+147 words/150 words
 
@@
 Partially observed confounder data are a major challenge for the statistical analysis of electronic health records (EHR).
-While analytic approaches exist (e.g., multiple imputation), assumption on underlying missingness patterns and mechanisms have to hold.
+While analytic approaches exist (e.g., multiple imputation), assumptions on underlying missingness patterns and mechanisms have to hold.
 We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.
@@
 
-We developed the smdi (structural missing data diagnostics) R package based on structural assumptions of common missing data mechanisms in real-world data and results of a previous simulation study using a US EHR-claims database linkage.
+We developed the smdi (structural missing data diagnostics) R package based on results of a previous simulation study which considered structural assumptions of common missing data mechanisms in real-world data.
 
@@
 
-smdi enables users to run principled missing data investigations on partially observed confounders including descriptive analyses and diagnostics on potentially underlying missingness patterns and mechanisms.
+smdi enables users to run principled missing data investigations on partially observed confounders and implements functions to visualize, describe and infer potential missingness patterns and mechanisms based on available data.
 
@@
 
-smdi can give valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.
-smdi is freely available seamlessly integrates to the ecosystem of R packages for healthcare database analytics.
+The smdi R package is freely available on CRAN and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can give valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.
 
@@
 
--   ...
+As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records).
+Hence, confounders and prognostic factors, which need to be balanced between treatment groups in real-world evidence studies, are usually not available for all patients at all necessary time points which leads to missing data.
+If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to be missing, this can lead to bias in the resulting effect estimates for the studied treatments if common missing data analysis methods like complete case analysis or imputation are used.
+Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for common analytic methods hold.
+Here, we present the smdi R package, which enables researchers to perform such principled missing data investigations on partially observed confounders.
+The smdi package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on available data.
 :::
@@
 Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
-These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
+These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
 However, these methods have so far been only described and tested in isolation from each other and no principled approach exists.
@@
 
-In light of these limitations, we have recently developed and evaluated a principled approach with multiple missing data diagnostics in an US EHR-claims database linkage [@weberpals2023].
+In light of these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023].
 The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data (e.g., missing data imputation).
@@
 
-To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
+To streamline and ease the implementation of these routine missing data diagnostics for POC data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
 
@@
 The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`.
-To ensure the quality and robustness of the package, we implemented comprehensive unit tests with a coverage of 95.81% and established automated daily R CMD checks [@wickham2023r] via continuous integration and deployment.
+To ensure the quality and robustness of the package, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment.
 Additional resources such as documentation and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.
@@
 
-The workflow is generally categorized into descriptives/pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms.
-In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and how these can be interpreted.
+The workflow is generally categorized into descriptives, pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms.
+In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and how these results can be interpreted.
 Examples are illustrated using a simulated EHR dataset that is part of the package (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
+
 For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent the variables relevant for the study, i.e., exposure, outcome, fully observed covariates and the POCs.
@@
 To quantify such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone].
-The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs along with a corresponding range of minimum and maximum AMSDs observed.
+The function returns an *asmd* object which displays an aggregated summary of the average or median ASMD along with a corresponding range of minimum and maximum AMSDs for each POC, respectivley.
 The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC.
-For a graphical visualization of this, the function returns a plot illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2].
+For a graphical visualization of this, the function returns a plot for each POC illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2].
 
 The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test.
-Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test for each POC, which means that `smdi_hotelling()` returns a test statistic and p-value for each each POC.
-In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is MCAR.
+Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test, which means that `smdi_hotelling()` returns a test statistic and p-value for each POC.
+In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is (globally) MCAR.
 
@@
 The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
-This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the POC given observed covariates as the predictors.
+This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors.
 If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism.
@@
 In case of inflated AUC values (\>0.9), the function prompts a message to the user reporting the most important predictor.
-If another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to exclude this POC and run the diagnostics for each POC independently.
+If in such a scenario another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to run the diagnostics for each POC independently rather than jointly.
 
@@
 Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes.
-Besides the the regression type, a user needs to specify the column name that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be exponentiated or not.
+Besides the the regression type (`model` parameter), a user needs to specify the column name containing the outcome using the `form_lhs` parameter (e.g., `Surv(eventtime, status)` in case of a Cox model).
 The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.
@@
 
-The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and the global Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**).
+The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**).
 The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.
-
-{{< pagebreak >}}
 
@@
 We developed `smdi` to streamline routine diagnostics of missing data.
-By cross-checking the resulting diagnostic parameters to expected estimates ( @fig-results, \[@weberpals2023\]), the diagnostics can provide valuable insights into underlying missingness patterns and mechanisms and help elucidate if analytic approaches such as imputation analyses are viable options.
+By cross-checking the resulting diagnostic parameters to expected estimates ( @fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness patterns and mechanisms and help elucidate if analytic approaches such as imputation analyses are viable options.
 
 The package also comes with limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data.
-Hence, it's important that diagnostic results are complemented with substantive expert knowledge to factor in how covariates are measured in routine care processes and contextualize potential reasons for missingness.
+Hence, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care processes and contextualize potential reasons for missingness.
 
@@
 
-The smdi package is a powerful and convenient tool to implement and carry out principled missing data diagnostics in RWE studies and improve the robustness of studies involving POCs by helping elucidate if certain analytic approaches are viable options for a given dataset.
-Through its design it seamlessly integrates with other R packages commonly used for clinical reporting [@gtsummary; @gt; @mice; @naniar; @ggplot2].
+The smdi package is a powerful and convenient tool to implement and carry out principled routine missing data diagnostics in RWE studies.
+This will improve the robustness of studies involving POCs by helping elucidate if certain analytic approaches are viable options for a given dataset.
 

### 2023-08-07 README.Rmd
modified readme with paper title
minor modification
version before first co-author review round

@@
 ---
+
+# smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies
 
@@
 -   .Rprofile - defines paths and activates `renv`, options for Posit R package manager
+-   manuscript - manuscript, bibliography, cover letter and meta files to render to .docx format
 -   tables - main and supplementary tables (R objects and .docx format)
@@
 
-```{r}
+```{r, echo = FALSE}
 fs::dir_tree(recurse = 1)

### 2023-08-07 README.Rmd
modified readme

@@
 
--   .Rprofile - defines paths and activates `renv`, options for Posit R package manager
 -   manuscript - manuscript, bibliography, cover letter and meta files to render to .docx format
@@
 -   renv/renv.lock - `renv` directories to manage R dependencies and versions used in this simulation
+-   .Rprofile - defines paths and activates `renv`, options for Posit R package manager
 

### 2023-08-21 manuscript/manuscript.qmd
adeed DT comments

@@
 
-We developed the smdi (structural missing data diagnostics) R package based on results of a previous simulation study which considered structural assumptions of common missing data mechanisms in real-world data.
+We developed the smdi (structural missing data investigations) R package based on results of a previous simulation study which considered structural assumptions of common missing data mechanisms in EHR.
 
@@
 Hence, confounders and prognostic factors, which need to be balanced between treatment groups in real-world evidence studies, are usually not available for all patients at all necessary time points which leads to missing data.
-If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to be missing, this can lead to bias in the resulting effect estimates for the studied treatments if common missing data analysis methods like complete case analysis or imputation are used.
+If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to have missing data, this can lead to bias in the resulting effect estimates for the studied treatments if common missing data analysis methods like complete case analysis or imputation are used.
 Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for common analytic methods hold.
@@
 
-Missing data are ubiquitous in real-world databases and may lead to bias if not handled appropriately.
+Missing data are ubiquitous in EHR and may lead to bias if not handled appropriately.
 We developed `smdi` to streamline routine diagnostics of missing data.
-By cross-checking the resulting diagnostic parameters to expected estimates ( @fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness patterns and mechanisms and help elucidate if analytic approaches such as imputation analyses are viable options.
+By cross-checking the resulting diagnostic parameters to expected estimates (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness patterns and mechanisms and help elucidate if analytic approaches such as imputation analyses are viable options.
 

### 2023-08-21 manuscript/manuscript.qmd
implemented HL edits

@@
 
-We developed the smdi (structural missing data investigations) R package based on results of a previous simulation study which considered structural assumptions of common missing data mechanisms in EHR.
+We developed the `smdi` (structural missing data investigations) R package based on results of a previous simulation study which considered structural assumptions of common missing data mechanisms in EHR.
 
@@
 
-smdi enables users to run principled missing data investigations on partially observed confounders and implements functions to visualize, describe and infer potential missingness patterns and mechanisms based on available data.
+`smdi` enables users to run principled missing data investigations on partially observed confounders and implements functions to visualize, describe and infer potential missingness patterns and mechanisms based on available data.
 
@@
 
-The smdi R package is freely available on CRAN and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can give valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.
+The `smdi` R package is freely available on CRAN and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can give valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.
 
@@
 As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records).
-Hence, confounders and prognostic factors, which need to be balanced between treatment groups in real-world evidence studies, are usually not available for all patients at all necessary time points which leads to missing data.
+Hence, confounders and prognostic factors, which need to be balanced between treatment groups to infer causal treatment effects, are usually not available for all patients at all necessary time points which leads to missing data.
 If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to have missing data, this can lead to bias in the resulting effect estimates for the studied treatments if common missing data analysis methods like complete case analysis or imputation are used.
 Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for common analytic methods hold.
-Here, we present the smdi R package, which enables researchers to perform such principled missing data investigations on partially observed confounders.
-The smdi package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on available data.
+Here, we present the `smdi` R package, which enables researchers to perform such principled missing data investigations on partially observed confounders.
+The `smdi` package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on available data.
 :::
@@
 If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism.
-In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism.
+In case of values close 0.5, this would indicate a random prediction and translate to a potential MCAR or MNAR mechanism.
 

### 2023-08-22 manuscript/manuscript.qmd
modified language on group 2 diagnostic model; abbreviate missingness mechanisms in box

@@
 
--   **MCAR**: The missingness does not depend on any other observed or unobserved covariate(s).
+-   **Missing completely at random (MCAR)**: The missingness does not depend on any other observed or unobserved covariate(s).
 
--   **MAR**: The missingness depends and can be explained by other observed covariates.
+-   **Missing at random (MAR)**: The missingness depends and can be explained by other observed covariates.
 
--   **MNAR**: The missingness depends on unoberserved covariate(s).
+-   **Missing not at random (MNAR)**: The missingness depends on unoberserved covariate(s).
 For example, the missingness may be explained by other covariate(s) which is/are not observed in the underlying dataset (MNAR~unmeasured~).
@@
 In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and how these results can be interpreted.
-Examples are illustrated using a simulated EHR dataset that is part of the package (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
+Examples are illustrated using a synthetic EHR dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
 
@@
 If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism.
-In case of values close 0.5, this would indicate a random prediction and translate to a potential MCAR or MNAR mechanism.
+In case of values close 0.5, this would indicate the model is unable to discriminate missing vs.
+observed values based on available data; this could be due to a mechanism that is close to MCAR or one where the missingness is associated with unmeasured data (MNAR).
 
@@
 The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**).
+By cross-checking the resulting diagnostic parameters to expected estimates (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness mechanisms and thereby help elucidate if analytic approaches such as imputation analyses are viable options.
+
 The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.
@@
 
-Missing data are ubiquitous in EHR and may lead to bias if not handled appropriately.
-We developed `smdi` to streamline routine diagnostics of missing data.
-By cross-checking the resulting diagnostic parameters to expected estimates (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness patterns and mechanisms and help elucidate if analytic approaches such as imputation analyses are viable options.
+Missing data are ubiquitous in EHR and may lead to bias if not handled appropriately. We developed `smdi` to streamline routine diagnostics of missing data.
 

### 2023-08-29 manuscript/manuscript.qmd
implemented JC comments

@@
 Partially observed confounder data are a major challenge for the statistical analysis of electronic health records (EHR).
-While analytic approaches exist (e.g., multiple imputation), assumptions on underlying missingness patterns and mechanisms have to hold.
+While analytic approaches exist (e.g., multiple imputation), assumptions on underlying missingness patterns and mechanisms must hold.
 We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.
@@
 
-`smdi` enables users to run principled missing data investigations on partially observed confounders and implements functions to visualize, describe and infer potential missingness patterns and mechanisms based on available data.
+`smdi` enables users to run principled missing data investigations on partially observed confounders and implements functions to visualize, describe, and infer potential missingness patterns and mechanisms based on available data.
 
@@
 
-Electronic health records (EHR) are increasingly used to conduct real-world evidence (RWE) studies to complement evidence coming from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
-Due to their detailed capture of clinical parameters such as vital signs, lab measurements, physician assessments and lifestyle factors, EHRs can significantly improve the ability to control for confounding and imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
-However, such prognostic factors are often just partially observed which challenges the statistical analysis of the data and can result in severe bias when predicting or estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
+Electronic health records (EHR) are increasingly used to conduct real-world evidence (RWE) studies to complement evidence from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
+Due to their detailed capture of clinical parameters such as vital signs, lab measurements, physician assessments, and lifestyle factors, EHRs can significantly improve the ability to control for imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
+However, such prognostic factors are often partially observed which challenges the statistical analysis and can result in severe bias when predicting or estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
 
-In order to make an informed decision about the most appropriate analytic approach, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019].
-Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
+In order to make an informed decision about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019].
+Usually these are not known for a given RWE study, but general guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
 These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
@@
 
-In light of these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023].
+Given these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023].
 The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data (e.g., missing data imputation).
@@
 
-To streamline and ease the implementation of these routine missing data diagnostics for POC data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
+To streamline implementation of routine missing data diagnostics for POC data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
 
@@
 The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`.
-To ensure the quality and robustness of the package, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment.
-Additional resources such as documentation and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.
+To ensure quality, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment.
+Additional resources such as documentation and vignettes are provided on the package website: <https://janickweberpals.gitlab-pages.partners.org/smdi>.
 
@@
 
-The workflow is generally categorized into descriptives, pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms.
-In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and how these results can be interpreted.
-Examples are illustrated using a synthetic EHR dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
+The workflow is generally categorized into descriptives, pattern diagnostics, and inferential diagnostics on potentially underlying missingness mechanisms.
+In this section, we cover the principles behind the main package functions, a selection of parameters users can specify, the returned results and how these results can be interpreted.
+Examples are illustrated using a synthetic dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
 
-For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent the variables relevant for the study, i.e., exposure, outcome, fully observed covariates and the POCs.
-Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions.
+For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent relevant variables, i.e., exposure, outcome, fully observed covariates and the POCs.
+Any non-informative columns, such as patient identifiers, date columns, or zip codes should be dropped from the dataframe before calling the functions.
 Throughout all functions, the user has the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter).
@@
 The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC.
-For a graphical visualization of this, the function returns a plot for each POC illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2].
+The function returns a plot for each POC illustrating the ASMD for each compared patient characteristic as a graphical visualization (example @fig-examples c) [@ggplot2].
 
-The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test.
+The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining differences in patient characteristics as a formal statistical hypothesis test.
 Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test, which means that `smdi_hotelling()` returns a test statistic and p-value for each POC.
@@
 
-The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
+The group 2 diagnostics assess the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
 This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors.
@@
 
-The third group diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the POC and the outcome under study.
-The function will compute both a univariate model and a model adjusted for all other covariates included in the dataset.
+The third group of diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the POC and the outcome under study.
+The function computes both a univariate model and a model adjusted for all other covariates in the dataset.
 In preceding simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms [@weberpals2023].
@@
 
-Finally, the `smdi_diagnose()` function enables a user to compute all three group diagnostics with just one function call.
+Finally, the `smdi_diagnose()` function enables users to compute all three group diagnostics with a single function call.
 
@@
   data = smdi_data,
-  covar = NULL, # NULL includes all covariates with at least one NA
   model = "cox",
   form_lhs = "Surv(eventtime, status)",
-  n_cores = 3 # number of cores to parallelize computations on
+  n_cores = 3
   ) %>% 
@@
 
-The package also comes with limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data.
-Hence, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care processes and contextualize potential reasons for missingness.
+The package also has limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data.
+Hence, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care and contextualize potential reasons for missingness.
 
@@
 
-The smdi package is a powerful and convenient tool to implement and carry out principled routine missing data diagnostics in RWE studies.
-This will improve the robustness of studies involving POCs by helping elucidate if certain analytic approaches are viable options for a given dataset.
+The smdi package is a powerful and convenient tool to implement principled routine missing data diagnostics in RWE studies.
+This will improve the robustness of studies involving POCs by elucidating if certain analytic approaches are viable for a given dataset.
 

### 2023-08-29 manuscript/manuscript.qmd
shortened words to meet journal word count

@@
 
-[**Word count:**]{.underline} 2,070 words/2,000 words
+[**Word count:**]{.underline} 1,969 words/2,000 words
 
@@
 
-147 words/150 words
+145 words/150 words
 
@@
 Electronic health records (EHR) are increasingly used to conduct real-world evidence (RWE) studies to complement evidence from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
-Due to their detailed capture of clinical parameters such as vital signs, lab measurements, physician assessments, and lifestyle factors, EHRs can significantly improve the ability to control for imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
-However, such prognostic factors are often partially observed which challenges the statistical analysis and can result in severe bias when predicting or estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
+Due to their detailed capture of clinical parameters, EHRs can significantly improve the ability to control for imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
+However, such prognostic factors are often partially observed which challenges the statistical analysis and can result in severe bias when estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
 
-In order to make an informed decision about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019].
-Usually these are not known for a given RWE study, but general guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
+In order to make informed decisions about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019].
+Usually these are not known for a given RWE study, but guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
 These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
@@
 Given these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023].
-The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data (e.g., missing data imputation).
+The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data.
 
@@
 The workflow is generally categorized into descriptives, pattern diagnostics, and inferential diagnostics on potentially underlying missingness mechanisms.
-In this section, we cover the principles behind the main package functions, a selection of parameters users can specify, the returned results and how these results can be interpreted.
-Examples are illustrated using a synthetic dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
+In this section, we cover the principles behind the main package functions, a selection of parameters users can specify, the returned results and how these can be interpreted.
+Examples are illustrated using a synthetic dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details: <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
 
@@
 Any non-informative columns, such as patient identifiers, date columns, or zip codes should be dropped from the dataframe before calling the functions.
-Throughout all functions, the user has the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter).
+Throughout all functions, users have the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter).
 If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.
@@
 As a first step to explore the missingness in new datasets, the `smdi` package provides a few basic functions to describe and summarize missingness across all covariates.
-The `smdi_summarize()` function takes the dataframe as input and returns the amount and proportion of missing observations, which can also be stratified by a grouping variable (e.g., by an exposure or outcome variable).
+The `smdi_summarize()` function returns the amount and proportion of missing observations, which can also be stratified by a grouping variable.
 The `smdi_vis()` function returns a corresponding bar chart plot (example @fig-examples a).
@@
 To quantify such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone].
-The function returns an *asmd* object which displays an aggregated summary of the average or median ASMD along with a corresponding range of minimum and maximum AMSDs for each POC, respectivley.
-The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC.
-The function returns a plot for each POC illustrating the ASMD for each compared patient characteristic as a graphical visualization (example @fig-examples c) [@ggplot2].
+The function returns an *asmd* object which displays an aggregated summary of the average or median ASMD along with a corresponding range of minimum and maximum AMSDs for each POC, respectively.
+The object also returns detailed 'Table 1's' and plots [@ggplot2] for each POC displaying the distributions of observed covariates and resulting ASMDs between patients with and without an observed value for the POC (example @fig-examples c).
 
@@
 
-The group 2 diagnostics assess the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
+Group 2 diagnostics assess the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
 This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors.
@@
 The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC.
-The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio can be chosen by the user (`train_test_ratio` parameter).
+The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio can be specified (`train_test_ratio` parameter).
 The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d).
@@
 In preceding simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms [@weberpals2023].
-As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC.
+As expected, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC.
 Under MAR, given that missingness can be explained by observed covariates, a spurious association in the univariate model disappeared after adjustment.
@@
 Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes.
-Besides the the regression type (`model` parameter), a user needs to specify the column name containing the outcome using the `form_lhs` parameter (e.g., `Surv(eventtime, status)` in case of a Cox model).
+Besides the the regression type (`model` parameter), users need to specify the column containing the outcome using the `form_lhs` parameter (e.g., `Surv(eventtime, status)` in case of a Cox model).
 The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.
@@
   n_cores = 3
-  ) %>% 
-  smdi_style_gt()
+  )
 ```
@@
 The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**).
-By cross-checking the resulting diagnostic parameters to expected estimates (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness mechanisms and thereby help elucidate if analytic approaches such as imputation analyses are viable options.
+By cross-checking all resulting diagnostic parameters to expected estimates (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness mechanisms and thereby help elucidate if analytic approaches such as imputation analyses are viable options.
+
+```{=html}
+<!--
 
@@
 
+-->
+```
 # Discussion

### 2023-09-05 manuscript/manuscript.qmd
incorporated WL comments; fine-tuned FDA affiliations by specific department as per simulation manuscript comments

@@
 Connolly^5^, Kimberly J.
-Dandreo^5^, Fang Tian^4^, Wei Liu^4^, Jie Li^4^, José J.
-Hernández-Muñoz^4^, Robert J.
+Dandreo^5^, Fang Tian^6^, Wei Liu^6^, Jie Li^6^, José J.
+Hernández-Muñoz^6^, Robert J.
 Glynn^1^, Rishi J.
@@
 
-^1^Division of Pharmacoepidemiology and Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA
+^1^ Division of Pharmacoepidemiology and Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA
 
-^2^Department of Population Health Sciences, Duke University School of Medicine, Durham, NC
+^2^ Department of Population Health Sciences, Duke University School of Medicine, Durham, NC
 
-^3^Biostatistics Division, Kaiser Permanente Washington Health Research Institute, Seattle, WA
+^3^ Biostatistics Division, Kaiser Permanente Washington Health Research Institute, Seattle, WA
 
-^4^Center for Drug Evaluation and Research, Food and Drug Administration, Silver Spring, MD
+^4^ Office of Biostatistics, Center for Drug Evaluation and Research, US Food and Drug Administration, Silver Spring, MD
 
-^5^Department of Population Medicine, Harvard Medical School and Harvard Pilgrim Health Care Institute, Boston, MA
+^5^ Department of Population Medicine, Harvard Medical School and Harvard Pilgrim Health Care Institute, Boston, MA
+
+^6^ Office of Surveillance and Epidemiology, Center for Drug Evaluation and Research, US Food and Drug Administration, Silver Spring, MD
 
@@
 
-[**Word count:**]{.underline} 1,969 words/2,000 words
+[**Word count:**]{.underline} 1,970 words/2,000 words
 
@@
 
-`smdi` enables users to run principled missing data investigations on partially observed confounders and implements functions to visualize, describe, and infer potential missingness patterns and mechanisms based on available data.
+`smdi` enables users to run principled missing data investigations on partially observed confounders and implement functions to visualize, describe, and infer potential missingness patterns and mechanisms based on observed data.
 
@@
 Here, we present the `smdi` R package, which enables researchers to perform such principled missing data investigations on partially observed confounders.
-The `smdi` package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on available data.
+The `smdi` package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on observed data.
 :::
@@
 In this section, we cover the principles behind the main package functions, a selection of parameters users can specify, the returned results and how these can be interpreted.
-Examples are illustrated using a synthetic dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details: <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
+Examples are illustrated using a synthetic dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several baseline confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details: <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
 

### 2023-09-06 manuscript/manuscript.qmd
implemented FT comments

@@
 
-[**Word count:**]{.underline} 1,970 words/2,000 words
+[**Word count:**]{.underline} 1,983 words/2,000 words
 
@@
 
-145 words/150 words
+150 words/150 words
 
@@
 
-Partially observed confounder data are a major challenge for the statistical analysis of electronic health records (EHR).
-While analytic approaches exist (e.g., multiple imputation), assumptions on underlying missingness patterns and mechanisms must hold.
+Partially observed confounder data pose a major challenge when conducting causal inference and statistical analyses using electronic health records (EHR).
+While analytic approaches such as multiple imputation are available, assumptions on underlying missingness patterns and mechanisms must hold.
 We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.
@@
 
-The `smdi` R package is freely available on CRAN and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can give valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.
+The `smdi` R package is freely available on CRAN and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can provide valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.
 
@@
 
-As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records).
-Hence, confounders and prognostic factors, which need to be balanced between treatment groups to infer causal treatment effects, are usually not available for all patients at all necessary time points which leads to missing data.
-If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to have missing data, this can lead to bias in the resulting effect estimates for the studied treatments if common missing data analysis methods like complete case analysis or imputation are used.
-Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for common analytic methods hold.
-Here, we present the `smdi` R package, which enables researchers to perform such principled missing data investigations on partially observed confounders.
+As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records) purposes.
+Hence, confounders and prognostic factors, which need to be balanced between treatment groups to infer causal treatment effects, are often unavailable for all patients at all necessary time points, resulting in missing data.
+If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to have missing data, it can introduce bias into the resulting effect estimates for the studied treatments when common missing data analysis methods like complete case analysis or imputation are applied.
+Therefore, it is of utmost importance to investigate potential missingness patterns and mechanisms to determine if assumptions for common analytic methods hold.
+Here, we present the `smdi` R package, which enables researchers to conduct such principled missing data investigations on partially observed confounders.
 The `smdi` package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on observed data.
@@
 
-Electronic health records (EHR) are increasingly used to conduct real-world evidence (RWE) studies to complement evidence from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
-Due to their detailed capture of clinical parameters, EHRs can significantly improve the ability to control for imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
-However, such prognostic factors are often partially observed which challenges the statistical analysis and can result in severe bias when estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
+Electronic health records (EHR) are increasingly used to generate real-world evidence (RWE) to complement data from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
+Due to their comprehensive capture of clinical parameters, EHRs can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
+However, such prognostic factors are often partially observed, posing a challenge to statistical analysis and potentially leading to severe bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
 
@@
 These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
-However, these methods have so far been only described and tested in isolation from each other and no principled approach exists.
-In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently not often performed [@carroll2020; @wood2004; @harel2012].
+However, these methods have so far only been described and tested in isolation from each other and no principled approach exists.
+In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently infrequently performed [@carroll2020; @wood2004; @harel2012].
 
@@
 
-Given these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023].
-The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data.
+Considering these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023].
+The results of this large-scale study revealed that the combination of these diagnostics effectively characterized missing data mechanisms and provided helpful guidance for the choice of appropriate analytic methods to handle POC data.
 
@@
 
-To streamline implementation of routine missing data diagnostics for POC data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
+To streamline the implementation of routine missing data diagnostics for POC data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].
 
@@
 
-Missing data are ubiquitous in EHR and may lead to bias if not handled appropriately. We developed `smdi` to streamline routine diagnostics of missing data.
+Missing data are ubiquitous in RWE studies involving EHR and may introduce bias if not handled appropriately.
+To address this issue, we developed the `smdi` R package to streamline routine diagnostics of missing data.
 
-The package also has limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data.
-Hence, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care and contextualize potential reasons for missingness.
+The package does come with certain limitations, such that the true underlying mechanism causing the missing data can never be inferred with absolute certainty from the observed data.
+Therefore, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care and contextualize potential reasons for missingness.
 
@@
 
-The smdi package is a powerful and convenient tool to implement principled routine missing data diagnostics in RWE studies.
+The smdi R package is a powerful and convenient tool to implement principled routine missing data diagnostics in RWE studies.
 This will improve the robustness of studies involving POCs by elucidating if certain analytic approaches are viable for a given dataset.

### 2023-09-09 manuscript/manuscript.qmd
incorporated PAS edits and comments

@@
 
-[**Word count:**]{.underline} 1,983 words/2,000 words
+[**Word count:**]{.underline} 1,997 words/2,000 words
 
@@
 
-As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records) purposes.
+Contrary to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records) purposes.
 Hence, confounders and prognostic factors, which need to be balanced between treatment groups to infer causal treatment effects, are often unavailable for all patients at all necessary time points, resulting in missing data.
-If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to have missing data, it can introduce bias into the resulting effect estimates for the studied treatments when common missing data analysis methods like complete case analysis or imputation are applied.
+If the underlying missingness mechanisms of such factors are not observed, for example, if patients with higher levels of a prognostic biomarker are more likely to have missing data for that biomarker, it can introduce bias into the resulting effect estimates for the studied treatments when common missing data analysis methods like complete case analysis or imputation are applied.
 Therefore, it is of utmost importance to investigate potential missingness patterns and mechanisms to determine if assumptions for common analytic methods hold.
@@
 These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
-However, these methods have so far only been described and tested in isolation from each other and no principled approach exists.
+However, these methods have so far only been described and tested in isolation from each other and no unified principled approach exists.
 In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently infrequently performed [@carroll2020; @wood2004; @harel2012].
@@
 The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC.
-The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio can be specified (`train_test_ratio` parameter).
+The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio, number of trees and CPU cores to parallelize over can be specified (`train_test_ratio, ntree` and `n_cores` parameter, respectively).
 The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d).
@@
 
-The smdi R package is a powerful and convenient tool to implement principled routine missing data diagnostics in RWE studies.
+The `smdi` R package is a powerful and convenient tool to implement principled routine missing data diagnostics in RWE studies.
 This will improve the robustness of studies involving POCs by elucidating if certain analytic approaches are viable for a given dataset.
@@
 
-`r table_cap(name = "tbl2", "Example output of the smdi_diagnose() function applied to the examplary smdi_data dataset.")`
+`r table_cap(name = "tbl2", "Example output of the smdi_diagnose() function applied to the examplary smdi_data dataset.
+In this example, ECOG performance score (ecog_cat) shows no imbalances in patient characteristics between patient with and without an observed value (absolute standardized mean difference [ASMD] 0.029, p[Hotelling] 0.783, group 1 diagnostic).
+Additionally missingness cannot be predicted well (AUC = 0.
+510, group 2 diagnostic) and no difference in the outcome can be observed between patients with and without ecog_cat (log HR -0.06 [95% CI -0.16; 0.03], group 3 diagnostic).
+Accordingly, the missingness diagnostics indicate that ECOG follows a missing completely at random missingness (MCAR) mechanism.
+Similarly, the EGFR (egfr_cat) and PD-L1 (pdl1_num) biomarker variables can be interpreted as following a missing at random (MAR) and missing not at random value (MNARvalue) mechanism.
+See also @fig-results.")`
 
@@
 #| echo: false
+
 smdi_diagnose(
@@
   form_lhs = "Surv(eventtime, status)",
-  n_cores = 3,
+  n_cores = 3
   ) %>% 
@@
 #| out-width: 110%
+
 knitr::include_graphics(here::here("figures", "Figure_1_workflow.jpg"))
@@
 #| label: fig-examples
-#| fig-cap: "Exemplary visual outputs of the a) smdi_vis(), b) gg_miss_upset(), c) smdi_asmd() and d) smdi_rf() functions, respectively."
+#| fig-cap: "Exemplary visual outputs of the a) smdi_vis(), b) gg_miss_upset(), c) smdi_asmd() and d) smdi_rf() functions, respectively.
+Sub-figure a) displays the proportion of missing observations for each partially observed covariate stratified by exposure.
+The upset plot in sub-figure b) demonstrates how a monotone missingness pattern between partially observed covariates can be visually inspected using a set visualization technique [@ruddle2022].
+Sub-figure c) illustrates absolute standardized mean differences (ASMD) in patient characteristics between patients with and without a value observed for the PD-L1 (pdl1_num) biomarker as a measure of imbalance.
+Sub-figure d) demonstrates the variable importance of fully observed covariates for predicting missingness in the partially oberserved ECOG performance score variable (ecog_cat)."
 #| echo: false
@@
 #| fig-dpi: 400
-#| fig-width: 15
-#| fig-height: 10
+#| fig-width: 22
+#| fig-height: 11
 
@@
 panel_a <- smdi_vis(data = smdi_data, strata = "exposure")
-panel_a <- panel_a + ggplot2::theme_bw(base_size = 15)
+panel_a <- panel_a + ggplot2::theme_bw(base_size = 18)
 
@@
 
-panel_b <- as.ggplot(smdi::gg_miss_upset(data = data_monotone))
+panel_b <- as.ggplot(smdi::gg_miss_upset(
+  data = data_monotone,
+  point.size = 3.2, 
+  text.scale = 2
+  )
+  ) +
+  labs(
+    title = "Upset plot iIllustrating missingness patterns among partially observed covariates"
+  ) +
+  theme(plot.title = element_text(hjust = 0.6, size = 20))
 
@@
 asmd <- smdi_asmd(data = smdi_data)
-panel_c <- asmd$pdl1_num$asmd_plot
+panel_c <- asmd$pdl1_num$asmd_plot + 
+  ggplot2::theme_bw(base_size = 18)
 
@@
 rf <- smdi_rf(data = smdi_data, n_cores = 3)
-panel_d <- rf$ecog_cat$rf_plot
+panel_d <- rf$ecog_cat$rf_plot +
+  ggplot2::theme_bw(base_size = 18)
 
@@
              ),
-  label_size = 18,
+  label_size = 20,
   label_colour = "black",
@@
 
-jpeg(here::here("figures", "Figure_2_visuals.jpg"), width = 16, height = 10, units = "in", res=400)
+jpeg(here::here("figures", "Figure_2_visuals.jpg"), width = 22, height = 11, units = "in", res=400)
 panel
@@
 #| echo: false
+#| fig-dpi: 400
+#| fig-width: 22
+#| fig-height: 11
 knitr::include_graphics(here::here("figures", "Figure_3_results.jpg"))

### 2023-09-11 manuscript/manuscript.qmd
implemented RJD comments

@@
 
-[**Word count:**]{.underline} 1,997 words/2,000 words
+[**Word count:**]{.underline} 1,986 words/2,000 words
 
@@
 
-Partially observed confounder data pose a major challenge when conducting causal inference and statistical analyses using electronic health records (EHR).
-While analytic approaches such as multiple imputation are available, assumptions on underlying missingness patterns and mechanisms must hold.
-We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.
+Partially observed confounder data pose a major challenge in statistical analyses aimed to inform causal inference using electronic health records (EHR).
+While analytic approaches such as imputation are available, assumptions on underlying missingness patterns and mechanisms must be verified.
+We aimed to develop a toolkit to streamline missing data diagnostics to guide choice of analytic approaches based on meeting necessary assumptions.
 
@@
 
-The `smdi` R package is freely available on CRAN and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can provide valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.
+The `smdi` R package is freely available on CRAN and can provide valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.
 
@@
 Hence, confounders and prognostic factors, which need to be balanced between treatment groups to infer causal treatment effects, are often unavailable for all patients at all necessary time points, resulting in missing data.
-If the underlying missingness mechanisms of such factors are not observed, for example, if patients with higher levels of a prognostic biomarker are more likely to have missing data for that biomarker, it can introduce bias into the resulting effect estimates for the studied treatments when common missing data analysis methods like complete case analysis or imputation are applied.
-Therefore, it is of utmost importance to investigate potential missingness patterns and mechanisms to determine if assumptions for common analytic methods hold.
+It is of utmost importance to investigate potential missingness patterns and mechanisms to determine if assumptions for common analytic methods such as complete case analysis or imputations hold to avoid bias in resulting treatment effect estimates.
 Here, we present the `smdi` R package, which enables researchers to conduct such principled missing data investigations on partially observed confounders.
@@
 Electronic health records (EHR) are increasingly used to generate real-world evidence (RWE) to complement data from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
-Due to their comprehensive capture of clinical parameters, EHRs can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
+Due to their comprehensive capture of clinical details, EHRs can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
 However, such prognostic factors are often partially observed, posing a challenge to statistical analysis and potentially leading to severe bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
@@
 In order to make informed decisions about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019].
-Usually these are not known for a given RWE study, but guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
+Existing guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
 These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021].
@@
 Considering these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023].
-The results of this large-scale study revealed that the combination of these diagnostics effectively characterized missing data mechanisms and provided helpful guidance for the choice of appropriate analytic methods to handle POC data.
+The results of this large-scale study revealed that the combination of these diagnostics effectively identified underlying mechanisms and provided helpful guidance for the choice of appropriate analytic methods to handle POC data.
 

### 2023-09-13 manuscript/manuscript.qmd
implemented RJG comments

@@
 
-[**Word count:**]{.underline} 1,986 words/2,000 words
+[**Word count:**]{.underline} 1,996 words/2,000 words
 
@@
 Hence, confounders and prognostic factors, which need to be balanced between treatment groups to infer causal treatment effects, are often unavailable for all patients at all necessary time points, resulting in missing data.
-It is of utmost importance to investigate potential missingness patterns and mechanisms to determine if assumptions for common analytic methods such as complete case analysis or imputations hold to avoid bias in resulting treatment effect estimates.
+It is important to investigate potential missingness patterns and mechanisms to determine if assumptions for common analytic methods such as complete case analysis or imputations hold to avoid bias in resulting treatment effect estimates.
 Here, we present the `smdi` R package, which enables researchers to conduct such principled missing data investigations on partially observed confounders.
@@
 Electronic health records (EHR) are increasingly used to generate real-world evidence (RWE) to complement data from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
-Due to their comprehensive capture of clinical details, EHRs can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
-However, such prognostic factors are often partially observed, posing a challenge to statistical analysis and potentially leading to severe bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
+Due to their capture of clinical details, EHRs can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
+However, such prognostic factors are often partially observed, posing a challenge to statistical analysis and potentially leading to bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
 
-In order to make informed decisions about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019].
+In order to inform decisions about the most appropriate analytic approach, it is useful to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019].
 Existing guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms.
@@
 However, these methods have so far only been described and tested in isolation from each other and no unified principled approach exists.
-In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently infrequently performed [@carroll2020; @wood2004; @harel2012].
+In addition, the practical implementation of such diagnostics is time-consuming and consequently infrequently performed [@carroll2020; @wood2004; @harel2012].
 
@@
 -   Monotone pattern: If Y~j~ is the j^th^ column in a dataset *Y*, a missing data pattern is said to be *monotone* if the variables Y~j~ can be ordered such that if Y~j~ is missing then all variables Y~k~ with *k \> j* are also missing.
-This can occur, for example, in longitudinal studies with drop-out or with clinical labs which are typically measured together as part of a lab panel (e.g., renal or liver panel).
+This can occur, for example, in longitudinal studies with drop-out.
 
@@
 For example, the missingness may be explained by other covariate(s) which is/are not observed in the underlying dataset (MNAR~unmeasured~).
-The missingness can also just dependent on the actual value of the partially observed covariate itself (MNAR~value~).
+The missingness can also just depend on the actual value of the partially observed covariate itself (MNAR~value~).
 :::
@@
 For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent relevant variables, i.e., exposure, outcome, fully observed covariates and the POCs.
-Any non-informative columns, such as patient identifiers, date columns, or zip codes should be dropped from the dataframe before calling the functions.
+Any non-informative columns, e.g., patient identifiers, should be dropped from the dataframe before calling the functions.
 Throughout all functions, users have the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter).
@@
 This function uses a set visualization technique to visually infer potential (non-)monotone patterns based on the number of intersecting missing observations across all POCs [@ruddle2022].
-For example, a monotone pattern would be visually evident if, for a set of two or more variables (e.g., lab1, lab2), all or the majority of missing records would be observed for both variables simultaneously (example @fig-examples b).
+For example, a monotone pattern could be visually evident if, for a set of two or more lab variables which are typically measured together as part of a lab panel (e.g.
+renal or liver panel), the missingness of one lab is indicative of the missingness in the other lab and hence all or the majority of combinations of cells are missing together (example @fig-examples b).
 
@@
 To quantify such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone].
-The function returns an *asmd* object which displays an aggregated summary of the average or median ASMD along with a corresponding range of minimum and maximum AMSDs for each POC, respectively.
+The function returns an *asmd* object which displays an aggregated summary of the average or median ASMD along with a corresponding range of minimum and maximum ASMDs for each POC, respectively.
 The object also returns detailed 'Table 1's' and plots [@ggplot2] for each POC displaying the distributions of observed covariates and resulting ASMDs between patients with and without an observed value for the POC (example @fig-examples c).
@@
 Group 2 diagnostics assess the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
-This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors.
-If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism.
+This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given exposure, outcome, follow-up time, and covariates plus missingness indicator for other POC as the predictors.
+If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR/against MCAR being the underlying missingness mechanism.
 In case of values close 0.5, this would indicate the model is unable to discriminate missing vs.
@@
 The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC.
-The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio, number of trees and CPU cores to parallelize over can be specified (`train_test_ratio, ntree` and `n_cores` parameter, respectively).
+The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio, number of trees and CPU cores to parallelize over can be specified (`train_test_ratio, ntree` and `n_cores` parameter, respectively) [@randomForest; @breiman2001].
 The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d).
@@
 In case of inflated AUC values (\>0.9), the function prompts a message to the user reporting the most important predictor.
-If in such a scenario another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to run the diagnostics for each POC independently rather than jointly.
+If in such a scenario missingness in another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to run the diagnostics for each POC independently rather than jointly.
 
@@
 The function computes both a univariate model and a model adjusted for all other covariates in the dataset.
-In preceding simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms [@weberpals2023].
+In simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms [@weberpals2023].
 As expected, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC.
-Under MAR, given that missingness can be explained by observed covariates, a spurious association in the univariate model disappeared after adjustment.
+Under MAR, given that missingness can be sufficiently explained by observed covariates, a spurious association in the univariate model disappeared after adjustment.
 If the missingness followed any MNAR mechanism, an association was observed regardless of adjustment.
@@
 Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes.
-Besides the the regression type (`model` parameter), users need to specify the column containing the outcome using the `form_lhs` parameter (e.g., `Surv(eventtime, status)` in case of a Cox model).
+Besides the regression type (`model` parameter), users need to specify the column containing the outcome using the `form_lhs` parameter (e.g., `Surv(eventtime, status)` in case of a Cox model).
 The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.

### 2023-10-16 manuscript/manuscript.qmd
Implemented SciComms and FDA review feedback, final version for initial submission

@@
 All R code, materials and depedencies can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>.
-The R package presented in this study can be downloaded from CRAN via `install.packages("smdi")` or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.
+The R package presented in this study can be downloaded from the comprehensive R archive network (CRAN) via `install.packages("smdi")` or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.
 
@@
 Electronic health records (EHR) are increasingly used to generate real-world evidence (RWE) to complement data from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
-Due to their capture of clinical details, EHRs can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
+Due to their capture of clinical details, EHR can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
 However, such prognostic factors are often partially observed, posing a challenge to statistical analysis and potentially leading to bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
@@
 
--   **Missing not at random (MNAR)**: The missingness depends on unoberserved covariate(s).
+-   **Missing not at random (MNAR)**: The missingness depends on unobserved covariate(s).
 For example, the missingness may be explained by other covariate(s) which is/are not observed in the underlying dataset (MNAR~unmeasured~).
@@
 The `smdi` R package was written in R language (version 4.2.1).
-The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`.
+The package is available on the comprehensive R archive network (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`.
 To ensure quality, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment.
@@
 Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test, which means that `smdi_hotelling()` returns a test statistic and p-value for each POC.
-In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is (globally) MCAR.
+In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data are (globally) MCAR.
 

### 2023-11-16 manuscript/manuscript.qmd
corrected word count, ready for initial submission
added author degrees on title page as per journal policy
implemented reviewer # 1 comments

@@
 
-**Authors**: Janick Weberpals^1^, Sudha R.
-Raman^2^, Pamela A.
-Shaw^3^, Hana Lee^4^, Bradley G.
-Hammill^2^, Sengwee Toh^5^, John G.
-Connolly^5^, Kimberly J.
-Dandreo^5^, Fang Tian^6^, Wei Liu^6^, Jie Li^6^, José J.
-Hernández-Muñoz^6^, Robert J.
-Glynn^1^, Rishi J.
-Desai^1^
+**Authors**: Janick Weberpals^1^, RPh, PhD, Sudha R.
+Raman^2^, PhD, Pamela A.
+Shaw^3^, PhD, MS, Hana Lee^4^, PhD, Bradley G.
+Hammill^2^, DrPH, Sengwee Toh^5^, ScD, John G.
+Connolly^5^, ScD, Kimberly J.
+Dandreo^5^, MS, Fang Tian^6^, PhD, Wei Liu^6^, PhD, Jie Li^6^, PhD, José J.
+Hernández-Muñoz^6^, PhD, Robert J.
+Glynn^1^, PhD, ScD, Rishi J.
+Desai^1^, PhD
 
@@
 
-[**Word count:**]{.underline} 1,996 words/2,000 words
+[**Word count:**]{.underline} 2,000 words/2,000 words
 
@@
 [**Analytical code and data sharing statement:**]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2.
-All R code, materials and depedencies can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>.
+All R code, materials and depedencies can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript> or <https://github.com/janickweberpals/smdi-manuscript>.
 The R package presented in this study can be downloaded from the comprehensive R archive network (CRAN) via `install.packages("smdi")` or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.
@@
 
-Electronic health records (EHR) are increasingly used to generate real-world evidence (RWE) to complement data from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
+Administrative health insurance claims databases have traditionally been the foundation for pharmacoepidemiologic studies.
+However, a notable drawback lies in their inability to capture detailed clinical nuances.
+To overcome this limitation, substantial initiatives are underway, for instance in the FDA Sentinel initiative [@Desai2021], linking claims databases and electronic health records (EHR) to generate real-world evidence (RWE) and complement data from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
 Due to their capture of clinical details, EHR can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
+At the moment, substantial efforts focusing on the linkage of claims databases and EHR are underway, for instance, in the FDA Sentinel Initiative [@Desai2021].
 However, such prognostic factors are often partially observed, posing a challenge to statistical analysis and potentially leading to bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
@@
 
-Considering these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023].
+Considering these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics [@weberpals2023] using a database linkage from the Mass General Brigham Research Patient Data Registry EHR in Boston [@nalichowski2006] linked with Medicare fee-for-service claims data [@cmsResdac].
 The results of this large-scale study revealed that the combination of these diagnostics effectively identified underlying mechanisms and provided helpful guidance for the choice of appropriate analytic methods to handle POC data.
@@
 In this section, we cover the principles behind the main package functions, a selection of parameters users can specify, the returned results and how these can be interpreted.
-Examples are illustrated using a synthetic dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several baseline confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details: <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
+Examples are illustrated using a synthetic dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several baseline confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following a MAR, MNAR and MCAR mechanism, respectively (more details: <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).
 
@@
 
+Applying group 1 diagnostics to the synthetic example dataset would indicate that the ECOG POC (median ASMD 0.03, min-max 0.00-0.07, p-value 0.78) does not show any differences in observed patient characteristics between patients with and without and observed value for ECOG which would give evidence for a MCAR mechanism (@fig-results bottom, Group 1 diagnostics - orange boxes) .
+Conversely, in the case of EGFR and PD-L1, the group 1 diagnostics display larger differences and hence may rather underlie a MAR or MNAR mechanism (@fig-examples c, @fig-results).
+
 #### Group 2 Diagnostics
@@
 
-```{=html}
-<!--
+@fig-results (Group 2 diagnostics - blue boxes), for example, illustrates the AUC values of the output of `smdi_rf()` when applied to the synthetic example dataset.
+Since the missingness of the EGFR POC follows a true MAR mechanism, it shows a an AUC of 0.63 which is expectedly meaningfully higher than what is observed for ECOG (0.51) and PD-L1 (0.52) which follow a true MCAR and MNAR mechanism, respectively.
 
-To tune the random forest model hyperparameters, the `smdi_rf()` function has a few arguments that can be specified by the user for optimization such as the number of trees to grow (`ntree`; default is 1,000 trees), the ratio of the split between train and test datasets (`train_test_ratio`; default is a 70/30 split) and the number of cores to parallelize the computation on (`n_cores`, default is 1), since this function can be very time consuming, especially with larger datasets.
-
--->
-```
 #### Group 3 Diagnostics
@@
 The function computes both a univariate model and a model adjusted for all other covariates in the dataset.
-In simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms [@weberpals2023].
-As expected, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC.
+In simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms (@fig-results top) \[@weberpals2023\].
+As expected, under a MCAR mechanism the simulation suggested no difference in the outcome between patients with and without a value for the POC.
 Under MAR, given that missingness can be sufficiently explained by observed covariates, a spurious association in the univariate model disappeared after adjustment.
@@
 
+Illustrating the application of `smdi_outcome()`with the example dataset, the derived logHR coefficients for the missingness indicators of the POCs (@fig-results bottom, Group 3 diagnostics - green boxes) conform with the results as expected by on our simulations [@weberpals2023].
+Here, EGFR displays no difference in the outcome after adjustment for fully observed covariates (logHR -0.01, 95% confidence intervals \[CI\] -0.10 to 0.09) indicating MAR, ECOG shows no difference in either unadjusted or adjusted model (logHR -0.06, -0.16 to 0.03) indicating MCAR, and PD-L1 exhibits differences in both models indicating MNAR.
+
+Demonstrating the utilization of smdi_outcome() using the example dataset, the derived logHR coefficients for the missingness indicators of the POCs (@fig-results bottom, Group 3 diagnostics - green boxes) align with the anticipated outcomes from our simulations [@weberpals2023].
+Specifically, EGFR manifests no discernible difference in the outcome after adjustment for fully observed covariates (logHR -0.01, 95% confidence intervals \[CI\] -0.10 to 0.09), suggesting a MAR mechanism.
+ECOG exhibits no distinction in either the unadjusted or adjusted model (logHR -0.06, -0.16 to 0.03), indicating MCAR.
+Conversely, PD-L1 showcases differences in the outcome in both models, suggesting an MNAR context.
+
 #### `smdi_diagnose()` to compute all three group diagnostics
 
-Finally, the `smdi_diagnose()` function enables users to compute all three group diagnostics with a single function call.
+Finally, the `smdi_diagnose()` function enables users to compute all of the above discussed group diagnostics within a single function call.
 
@@
 Therefore, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care and contextualize potential reasons for missingness.
+This collaborative approach allows for a contextualized understanding of potential causes for missing data in EHR.
 

### 2023-11-20 manuscript/supplement.qmd
started describing the methods in deep detail

--- 2023-07-12 manuscript/supplement.qmd
@@
 ---
-title: "Supplementary Appendix"
-subtitle: "smdi: An R package to perform structural missing data investigations"
+title: "Supplementary Material"
+subtitle: "smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies"
+
 echo: false
@@
     fig-cap-location: top
-    fig-align: center
     crossref:
-      fig-title: "Supplementary Figure"
-    toc: true
-    toc-depth: 2
-    number-sections: true
-    colorlinks: true
+      fig-title: '**Supplementary Figure**'
 
 editor: visual
-#bibliography: references.bib
-#csl: jamia.csl
+bibliography: references.bib
+csl: jamia.csl
 ---
@@
 #| include: false
-#| eval: true
-library(magick)
 library(cowplot)
 library(here)
-library(knitr)
+library(tidyverse)
+library(readxl)
+library(gt)
+library(captioner)
+library(smdi)
+library(ggimage)
+
+table_cap <- captioner(prefix = "Table")
+```
+
+**Authors**: Janick Weberpals^1^, RPh, PhD, Sudha R.
+Raman^2^, PhD, Pamela A.
+Shaw^3^, PhD, MS, Hana Lee^4^, PhD, Bradley G.
+Hammill^2^, DrPH, Sengwee Toh^5^, ScD, John G.
+Connolly^5^, ScD, Kimberly J.
+Dandreo^5^, MS, Fang Tian^6^, PhD, Wei Liu^6^, PhD, Jie Li^6^, PhD, José J.
+Hernández-Muñoz^6^, PhD, Robert J.
+Glynn^1^, PhD, ScD, Rishi J.
+Desai^1^, PhD
+
+[Author affiliations:]{.underline}
+
+^1^ Division of Pharmacoepidemiology and Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA
+
+^2^ Department of Population Health Sciences, Duke University School of Medicine, Durham, NC
+
+^3^ Biostatistics Division, Kaiser Permanente Washington Health Research Institute, Seattle, WA
+
+^4^ Office of Biostatistics, Center for Drug Evaluation and Research, US Food and Drug Administration, Silver Spring, MD
+
+^5^ Department of Population Medicine, Harvard Medical School and Harvard Pilgrim Health Care Institute, Boston, MA
+
+^6^ Office of Surveillance and Epidemiology, Center for Drug Evaluation and Research, US Food and Drug Administration, Silver Spring, MD
+
+{{< pagebreak >}}
+
+# Supplementary Methods
+
+## Missingness assumptions
+
+For the design and validation of the `smdi` R package functions, we employed comprehensive simulations and a real-world example [@weberpals2023a].
+These simulations followed structural assumptions on realistic missingness generating mechanisms that could be expected in electronic health record (EHR) data.
+The below directed acyclic diagrams (DAG) [@Moreno-Betancur2018; @sondhi2023a], where a directed edge represents a causal effect of one variable on another variable, outline and illustrate these assumptions along with a brief explanations and exemplary real-world EHR scenarios.
+Since the `smdi` package was developed with focus on partially observed confounders (as opposed to missingness in exposure or outcome), we simplified the below DAGs by leaving out the nodes for exposure and outcome.
+
+**Notation**: C = Fully observed confounder, POC = True values of the partially observed confounder, POCobs = Observed portion of the partially observed confounder, M = Missingness of POC with M=0 fully observed and M=1 fully missing, U = Unobserved confounder.
+
+### Missing completely at random (MCAR)
+
+The missingness (M) of the partially observed confounder (POC) is independent of any observed and unobserved confounders.
+That is, no edge directs to M and the missingness is completely at random (MCAR).
+
+[Real-world example:]{.underline} A machine breaks that is used to measure and analyze a certain biomarker.
+
+```{dot}
+//| label: fig-mcar
+//| fig-cap: "Directed acyclic graph displaying a missing completely at random (MCAR) scenario."
+//| fig-width: 4
+digraph {
+
+  rankdir = LR;
+    #Y [label="Outcome"]
+    #E [label="Exposure"]
+    #C [label="Observed confounder"]
+    #POC [label="Partially observed confounder"]
+    #POCobs [label="Observed portion of the partially observed confounder"]
+    #M [label="Missingness indicator"]
+
+    C -> POC
+    POC -> POCobs
+    M -> POCobs
+}
+```
+
+### Missing at random (MAR)
+
+The missingness (M) of the partially observed confounder (POC) can be explained by one or multiple fully observed confounders (C).
+That is, there is an edge between C to M and the missingness is at random (MAR).
+
+[Real-world example:]{.underline} Older patients are more likely to receive a certain biomarker test and age is a fully observed confounder which is measured for every patient in the dataset.
+
+```{dot}
+//| label: fig-mar
+//| fig-cap: "Directed acyclic graph displaying a missing completely at random scenario."
+//| fig-width: 4
+digraph {
+
+  rankdir = LR;
+
+    C -> POC
+    POC -> POCobs
+    C -> M
+    M -> POCobs
+}
+```
+
+### Missing not at random - unmeasured (MNAR~unmeasured~)
+
+The missingness (M) of the partially observed confounder (POC) can only be explained by one or multiple confounder(s) which are not observed in the dataset.
+
+[Real-world example:]{.underline} Patients with a certain performance status, which not observed in the dataset, are more likely to receive a lab test.
+
+```{dot}
+//| label: fig-mnarU
+//| fig-cap: "Directed acyclic graph displaying a missing not at random (unmeasured) scenario."
+//| fig-width: 4
+digraph {
+
+  rankdir = LR;
+
+    C -> POC
+    POC -> POCobs
+    M -> POCobs
+    U -> M
+    U -> POC
+}
+```
+
+### Missing not at random - value (MNAR~value~)
+
+The missingness (M) of the partially observed confounder (POC) depends on the true value of the partially observed confounder itself.
+
+[Real-world example:]{.underline} Patients with a history of normal biomarker values are systematically less likely to be tested again for the same biomarker.
+
+```{dot}
+//| label: fig-mnarV
+//| fig-cap: "Directed acyclic graph displaying a missing not at random (value) scenario."
+//| fig-width: 4
+digraph {
+
+  rankdir = LR;
+
+    C -> POC
+    POC -> POCobs
+    M -> POCobs
+    POC -> M
+}
 ```
@@
 
-# Supplementary material
+## Missingness characterization
+
+### Group 1 diagnostics
+
+The featured group 1 diagnostics in `smdi` focus on investigating potential differences in the distribution of characteristics between patients with and without an observed value for a partially observed confounder.
+To that end, different analytic approaches are leveraged which are described for each function below.
+
+-   `smdi_hotelling()`: This function computes a two-sample Hotelling's T-squared test for the difference in two multivariate means based on the Hotelling's T-Square test statistic [@hotelling1931; @Hotelling].
+It assesses whether there is a statistically significant difference between the means of two groups in multivariate data.
+This test is an extension of the univariate t-test that examines for two groups (i.e., patients with and without a value observed for the confounder of interest) with p variables (e.g., patient or disease characteristics) whether their mean vectors (p-dimensional vectors of means) are significantly different.
+To that end, the T\^2 statistic is derived by a [vector of covariate means](https://github.com/jmcurran/Hotelling/blob/090f742a631c256de36c1d714ebfa7bc7d90b234/R/hotelling.test.r#L84) (i.e, one element for each covariate) [@rpubsHotelling], which are represented as x bar 1 (vector of covariate means in group 1) and x bar 2 (vector of covariate means in group 2) in the equation [@pennStateHotelling] below (with n1 and n2 representing the sample sizes of group 1 and 2, respectively, and Sp representing the pooled variance-covariance matrix).
+To implement Hotelling's test in the `smdi` package, we built a wrapper around the `Hotelling::hotelling.test()` function [@Hotelling].
+
+$$T^2 = \mathbf{(\bar{x}_1 - \bar{x}_2)}^T\{\mathbf{S}_p(\frac{1}{n_1}+\frac{1}{n_2})\}^{-1} \mathbf{(\bar{x}_1 - \bar{x}_2)}$$
+
+-   `smdi_little()`: Little's test, as opposed to Hotelling's test, performs a single global test to evaluate the missing completely at random (MCAR) assumption [@little1988].
+It was developed since with an approach like Hotelling's test, there can be concerns regarding multiplicity due to the multiple testing of each variable.
+Little's test first [identifies a set of subgroups that share the same missing data patterns](https://github.com/njtierney/naniar/blob/master/R/mcar-test.R#L54C4-L54C4).
+Across these missing data patterns, it then tests for mean differences on every variable by comparing the observed means versus expected population means which are [estimated using a maximum likelihood (ML) expectation-maximization (EM) algorithm](https://github.com/njtierney/naniar/blob/master/R/mcar-test.R#L69) as implemented by the `norm::prelim.norm()` and `norm::em.norm()` functions in `R`.
+Under the assumption of MCAR, the test statistic ($d^2$) approximates a chi-square distribution and is derived by computing the sum of squared standardized differences between the subgroup means and the expected population means weighted by both the estimated variance-covariance matrix and the respective subgroup sizes [@mistyMCAR; @enders2022applied].
+To implement Little's test in the `smdi` package, we built a wrapper around the `naniar::mcar_test()` function [@naniar].
+
+$$d^2 = \sum_{j = 1} ^ {J} n_{j}(\hat{\mu}_{j} - \hat{\mu}{j ^ \text{(ML)}})^T \hat{\Sigma}_{j}^{-1} (\hat{\mu}_{j} - \hat{\mu}{j ^ \text{(ML)}})$$
+
+For both `smdi_hotelling()` and `smdi_little()`, a high test statistic and a low p-value would indicate differences in the groups compared.
+A limitation of both Hotelling's and Little's test is that they assume continuous variables following multivariate normality.
+Real-world data, however, are often of binary nature and to consider categorical data in the computation of these tests, [categorical data are one-hot encoded to binary dummy variables](https://gitlab.partners.org/janickweberpals/smdi/-/blob/main/R/smdi_hotelling.R?ref_type=heads#L89) before performing each test.
+
+-   `smdi_asmd()`:
 
@@
 
-# Supplementary Figures
-
-```{r}
-#| label: fig-GitHub-gui_a
-#| fig-cap: "Starting a new repository through the GitHub GUI."
-#| out-width: 115%
-
-include_graphics(here("figures", ""))
-```
-
-{{< pagebreak >}}
-
-```{r}
-#| label: fig-GitHub-gui_changes
-#| eval: false
-#| fig-cap: "Remote repositories like GitHub provide tools to visually track changes made to analytical code and corresponding output such as figures.
-This commit history illustrates details about the commit (e.g., commit SHA hash and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom)."
-#| out-width: 105%
-
-include_graphics(here("figures", "SFigure_4_commit_Github_gui.png"))
-```
-
 # References {.unnumbered}
 
-::: {#refs}
+::: refs
 :::
-
-{{< pagebreak >}}
-
-#  {.unnumbered}

### 2023-11-21 manuscript/supplement.qmd
progressed with reviewe #2 comments

@@
 
-{{< pagebreak >}}
+## Missingness characterization
 
-## Missingness characterization
+This sections provides extensive information and details on key statistical principles and methods employed in the `smdi` package.
+Details on all package functions can also be found in the corresponding documentation section of the `smdi` package website
+
+<https://janickweberpals.gitlab-pages.partners.org/smdi/reference/index.html>
+
+or accessed in `R` by executing the the function name preceded with a question mark, e.g.:
+
+```{r}
+#| eval: false
+#| echo: true
+?smdi_diagnose()
+```
 
@@
 The featured group 1 diagnostics in `smdi` focus on investigating potential differences in the distribution of characteristics between patients with and without an observed value for a partially observed confounder.
-To that end, different analytic approaches are leveraged which are described for each function below.
+To that end, different analytic approaches are leveraged which are described in detail for each function below.
 
@@
 This test is an extension of the univariate t-test that examines for two groups (i.e., patients with and without a value observed for the confounder of interest) with p variables (e.g., patient or disease characteristics) whether their mean vectors (p-dimensional vectors of means) are significantly different.
-To that end, the T\^2 statistic is derived by a [vector of covariate means](https://github.com/jmcurran/Hotelling/blob/090f742a631c256de36c1d714ebfa7bc7d90b234/R/hotelling.test.r#L84) (i.e, one element for each covariate) [@rpubsHotelling], which are represented as x bar 1 (vector of covariate means in group 1) and x bar 2 (vector of covariate means in group 2) in the equation [@pennStateHotelling] below (with n1 and n2 representing the sample sizes of group 1 and 2, respectively, and Sp representing the pooled variance-covariance matrix).
+To that end, the T\^2 statistic is derived by a [vector of covariate means](https://github.com/jmcurran/Hotelling/blob/090f742a631c256de36c1d714ebfa7bc7d90b234/R/hotelling.test.r#L84) (i.e, one element for each covariate) [@rpubsHotelling], which are represented as $\bar{x}_1$ (vector of covariate means in group 1) and $\bar{x}_2$ (vector of covariate means in group 2) in @eq-hotelling below (with $n_1$ and $n_2$ representing the sample sizes of group 1 and 2, respectively, and $S_p$ representing the pooled variance-covariance matrix) (@eq-hotelling adapted from [@pennStateHotelling]).
 To implement Hotelling's test in the `smdi` package, we built a wrapper around the `Hotelling::hotelling.test()` function [@Hotelling].
 
-$$T^2 = \mathbf{(\bar{x}_1 - \bar{x}_2)}^T\{\mathbf{S}_p(\frac{1}{n_1}+\frac{1}{n_2})\}^{-1} \mathbf{(\bar{x}_1 - \bar{x}_2)}$$
+$$T^2 = \mathbf{(\bar{x}_1 - \bar{x}_2)}^T\{\mathbf{S}_p(\frac{1}{n_1}+\frac{1}{n_2})\}^{-1} \mathbf{(\bar{x}_1 - \bar{x}_2)}$$ {#eq-hotelling}
 
@@
 It was developed since with an approach like Hotelling's test, there can be concerns regarding multiplicity due to the multiple testing of each variable.
-Little's test first [identifies a set of subgroups that share the same missing data patterns](https://github.com/njtierney/naniar/blob/master/R/mcar-test.R#L54C4-L54C4).
-Across these missing data patterns, it then tests for mean differences on every variable by comparing the observed means versus expected population means which are [estimated using a maximum likelihood (ML) expectation-maximization (EM) algorithm](https://github.com/njtierney/naniar/blob/master/R/mcar-test.R#L69) as implemented by the `norm::prelim.norm()` and `norm::em.norm()` functions in `R`.
-Under the assumption of MCAR, the test statistic ($d^2$) approximates a chi-square distribution and is derived by computing the sum of squared standardized differences between the subgroup means and the expected population means weighted by both the estimated variance-covariance matrix and the respective subgroup sizes [@mistyMCAR; @enders2022applied].
+Little's test first [identifies a set of subgroups that share the same missing data patterns $j$](https://github.com/njtierney/naniar/blob/master/R/mcar-test.R#L54C4-L54C4).
+Across these missing data patterns ($n_j$), it then tests for mean differences on every variable by comparing the observed means ($\hat{\mu}_{j}$) versus expected population means ($\hat{\mu}{j ^ \text{(ML)}}$) which are [estimated using a maximum likelihood (ML) expectation-maximization (EM) algorithm](https://github.com/njtierney/naniar/blob/master/R/mcar-test.R#L69) as implemented by the `norm::prelim.norm()` and `norm::em.norm()` functions in `R` (@eq-little adapted from [@enders2022applied]).
+Under the assumption of MCAR, the test statistic ($d^2$) approximates a chi-square distribution and is derived by computing the sum of squared standardized differences between the subgroup means and the expected population means weighted by both the estimated variance-covariance matrix (where $\hat{\Sigma}_{j}$ is the maximum likelihood estimate of the covariance matrix) and the respective subgroup sizes [@mistyMCAR; @enders2022applied].
 To implement Little's test in the `smdi` package, we built a wrapper around the `naniar::mcar_test()` function [@naniar].
 
-$$d^2 = \sum_{j = 1} ^ {J} n_{j}(\hat{\mu}_{j} - \hat{\mu}{j ^ \text{(ML)}})^T \hat{\Sigma}_{j}^{-1} (\hat{\mu}_{j} - \hat{\mu}{j ^ \text{(ML)}})$$
+$$d^2 = \sum_{j = 1} ^ {J} n_{j}(\hat{\mu}_{j} - \hat{\mu}{j ^ \text{(ML)}})^T \hat{\Sigma}_{j}^{-1} (\hat{\mu}_{j} - \hat{\mu}{j ^ \text{(ML)}})$$ {#eq-little}
 
@@
 Real-world data, however, are often of binary nature and to consider categorical data in the computation of these tests, [categorical data are one-hot encoded to binary dummy variables](https://gitlab.partners.org/janickweberpals/smdi/-/blob/main/R/smdi_hotelling.R?ref_type=heads#L89) before performing each test.
+In addition, these tests are sensitive to just small differences when study sizes are large which is why we recommend to perform these tests jointly with `smdi_asmd`.
 
--   `smdi_asmd()`:
+-   `smdi_asmd()`: The `smdi_asmd` functions computes the absolute standardized mean difference (ASMD) as a statistical measure for assessing dissimilarities in disease and patient characteristics between individuals with and without observed value for a specific partially observed confounder (POC) of interest.
+If the median/average ASMD is high, this may indicate imbalance in patient covariate distributions which may be indicative of the POC following a missing at random (MAR) mechanims, i.e.
+the missingness is explainable by other observed covariates.
+Similarly, no imbalance between observed covariates may be indicative that missingness cannot be explained with observed covariates and the underlying missingness mechanism may be completely at random (MCAR) or not at random (e.g., missingness is only associated with unobserved factors or through the POC itself).
+This methodological approach follows the same theory that is often applied to measure covariate balance between two groups before and after matching or weighting on propensity scores, which are balancing scores often used to reduce confounding in real-world evidence studies [@austin2011].
+While there isn't a universally established standard for the threshold of the ASMD indicating significant imbalance, an ASMD below 0.1 is often considered indicative of a negligible difference in the mean or prevalence of a covariate between two groups [@Normand2001; @austin2011].
+The advantage of ASMDs to assess the balance of patient characteristics between two groups with and without an observed value for the POC is that they are not influenced by sample size, do not come with many assumptions, are easy to interpret and applicable to various types of covariates.
+@eq-asmd-cont and @eq-asmd-binary illustrate how ASMDs are computed for continous and binary covariates with $\bar{x}$, $s^2$ and $\hat{p}$ indicating the mean, sample variance and prevalence of the binary variable of the covariate in patients with a missing ($M$) and an observed ($O$) value for the POC, respectively.
+In the `smdi` package, ASMDs are [extracted from tableone objects](https://github.com/kaz-yos/tableone/blob/master/R/modules-smd.R) using the `ExtractSmd` function [@tableone].
+
+$$ASMD_{continous} = |\frac{(\bar{x}_M - \bar{x}_O)}{\sqrt{\frac{s^2_M + s^2_O}{2}}}|$$ {#eq-asmd-cont}
+
+$$ASMD_{binary} = | \frac{(\hat{p_M} - \hat{p_O})}{\sqrt{\frac{\hat{p_M}(1-\hat{p_M}) + \hat{p_O}(1-\hat{p_O})}{2}}} |$$ {#eq-asmd-binary}
+
+### Group 2 diagnostics
+
+-   `smdi_rf()`: The Group 2 diagnostics, implemented in the `smdi_rf()` function, assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
+This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given exposure, outcome, follow-up time, and covariates plus missingness indicator for other POC as the predictors.
+We chose a random forest classification model for the purpose of this package due its many beneficial features within the context of data structures that are frequently encountered in routine healthcare databases:
+
+    -   Ability to implicitly model nonlinear and non-additive relationships between observed variables (i.e., higher order terms do not need to be explicitly specified).
+
+    -   Recursive partitioning models like random forests have been found to work particularly well with sparse tabular data, which is the typical data type that is used for real-world evidence studies [@SHWARTZZIV202284; @grinsztajn2022treebased].
+
+    -   Random Forests provide transparent feature importance measures, indicating the variables contributing significantly to predicting missingness.
+This aids in identifying key features driving the missingness mechanisms in the dataset.
+
+    -   Multiple other studies have also reported good result of this random forest-based approach [@sondhi2023a; @beaulieu-jones2018].
+
+The default training parameters of the random forest model include a train-test split of the data (default if 70% training and 30% testing) which can be changed using the `train_test_ratio` parameter and the number of trees to grow using `ntree` (default is 1000 decision trees since a higher number of trees typically give more stable results).
+Users can optionally select `tune = TRUE` which will perform a 5-fold cross validation and a random search for the optimal number of variables randomly sampled as candidates at each split (`mtry`).
+However, users should be aware that this may lead to longer computation times with larger datasets which is why the default is set to `FALSE`.
+In summary, the following parameters are part of this function:
+
+-   `data`: dataframe or tibble object with partially observed/missing variables
+
+-   `covar`: character covariate or covariate vector with partially observed variable/column name(s) to investigate.
+If NULL, the function automatically includes all columns with at least one missing observation and all remaining covariates will be used as predictors
+
+-   `train_test_ratio`: numeric vector to indicate the test/train split ratio, e.g. c(.7, .3) which is the default
+
+-   `set_seed`: seed for reproducibility, defaults to 42
+
+-   `ntree`: integer, number of trees (defaults to 1000 trees)
+
+-   `n_cores`: integer, if \>1, computations will be parallelized across the amount of cores specified in `n_cores`.
+This is important especially for larger datasets as the random forest training can be very time-consuming if done sequentially.
+
+### Group 3 diagnostics
+
+To evaluate the association of the misssing indicator variable of a POC ($M_{POC}$) and the studied outcome in Group 3 Diagnostics, conventional univariate and multivariate regression models which are found used in the vast majority of real-world evidence studies are employed.
+The selection of regression type needs to be chosen using the `model` parameter and is determined by the type of outcome that is studied.
+The possible outcome models include linear regression [@stats] for continuous outcomes (@eq-linear), logistic regression [@glmnet] for binary outcomes (@eq-logistic) and a Cox proportional hazards regression [@survival] for time-to-event outcomes (@eq-cox) with $X_j$ indicating other covariates included in the multivariate models.
+The `form_lhs` specifies the left-hand side of the outcome formula, which, in case of `model = "linear"` or `model = "logistic` just reflects the name of the column that contains the outcome and the form `Surv(time, status)` for `model ="cox"` indicating the time-to-event variable and the event status (0/1).
+The output of the functions automatically returns results for both the univariate and the multivariate models and a user has the choice to have the $\beta_1M_{POC}$ estimate exponentiated or not using the logical `exponentiated` parameter.
+
+$$Y = \beta_0 + \beta_1M_{POC} + ...
++ \beta_2X_j$$ {#eq-linear} $$\ln(\frac{p}{p-1}) = \beta_0 + \beta_1M_{POC} + ...
++ \beta_2X_j$$ {#eq-logistic} $$h(t) = h_0(t) e^{\sum{\beta_1M_{POC} + ...
++  \beta_2X_j}} $$ {#eq-cox}
 

### 2023-11-21 manuscript/manuscript.qmd
progressed with reviewe #2 comments
finished reviewer #2 review and finalized all responses, supplement and manuscript updates.Sending revised version now to  co-author review.

@@
 
-[**Word count:**]{.underline} 2,000 words/2,000 words
+[**Word count:**]{.underline} 2,514 words
 
-[**Tables:**]{.underline} 2/2
+[**Tables:**]{.underline} 2
 
-[**Figures:**]{.underline} 3/3
+[**Figures:**]{.underline} 3
 
-[**Supplementary material:**]{.underline} None
+[**Supplementary material:**]{.underline} Supplementary Methods
 
@@
 All R code, materials and depedencies can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript> or <https://github.com/janickweberpals/smdi-manuscript>.
-The R package presented in this study can be downloaded from the comprehensive R archive network (CRAN) via `install.packages("smdi")` or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.
+The R package presented in this study can be downloaded from the comprehensive R archive network (CRAN) via `install.packages("smdi")` (version 0.2.2 at time of manuscript submission) or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.
 
@@
 At the moment, substantial efforts focusing on the linkage of claims databases and EHR are underway, for instance, in the FDA Sentinel Initiative [@Desai2021].
-However, such prognostic factors are often partially observed, posing a challenge to statistical analysis and potentially leading to bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
+However, prognostic factors coming from EHR are often only partially observed, posing a challenge to the statistical analysis and potentially leading to bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].
 
@@
 
+Details on missingness assumptions, key statistical principles and further information on all functions can be found in the **Supplementary Methods** and in the documentation of each respective function which can be accessed in R by preceding the function name with a question mark, e.g.:
+
+```{r}
+#| eval: false
+#| echo: true
+?smdi_diagnose()
+```
+
 ### Descriptives and Pattern Diagnostics
@@
 renal or liver panel), the missingness of one lab is indicative of the missingness in the other lab and hence all or the majority of combinations of cells are missing together (example @fig-examples b).
+The `md.pattern()` function, a re-export of the mice package [@mice], provides a similar functionality and returns a matrix displaying the frequency of each observed missing data pattern.
 
@@
 @fig-results (Group 2 diagnostics - blue boxes), for example, illustrates the AUC values of the output of `smdi_rf()` when applied to the synthetic example dataset.
-Since the missingness of the EGFR POC follows a true MAR mechanism, it shows a an AUC of 0.63 which is expectedly meaningfully higher than what is observed for ECOG (0.51) and PD-L1 (0.52) which follow a true MCAR and MNAR mechanism, respectively.
+Since the missingness of the EGFR POC follows a true MAR mechanism, the resulting AUC of 0.63 is expectedly meaningfully higher than what is observed for ECOG (0.51) and PD-L1 (0.52) which follow a true MCAR and MNAR mechanism, respectively.
 
@@
 The function computes both a univariate model and a model adjusted for all other covariates in the dataset.
-In simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms (@fig-results top) \[@weberpals2023\].
+In simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms (@fig-results top) [@weberpals2023].
 As expected, under a MCAR mechanism the simulation suggested no difference in the outcome between patients with and without a value for the POC.
@@
 
-Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes.
-Besides the regression type (`model` parameter), users need to specify the column containing the outcome using the `form_lhs` parameter (e.g., `Surv(eventtime, status)` in case of a Cox model).
+`smdi_outcome()` supports multiple outcome regression types including linear regression (*lm* [@stats]) for continuous outcomes, Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes, and generalized linear regression models (*glm* [@stats]) for which the family of conditional distributions of the outcome can be selected using the `glm_family` parameter (the default is `binomial(link='logit')`).
+Besides the regression type (`model` parameter) and the `glm_family` (in case of a glm model), users need to specify the column containing the outcome using the `form_lhs` parameter (e.g., `Surv(eventtime, status)` in case of a Cox model).
 The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.
-
-Illustrating the application of `smdi_outcome()`with the example dataset, the derived logHR coefficients for the missingness indicators of the POCs (@fig-results bottom, Group 3 diagnostics - green boxes) conform with the results as expected by on our simulations [@weberpals2023].
-Here, EGFR displays no difference in the outcome after adjustment for fully observed covariates (logHR -0.01, 95% confidence intervals \[CI\] -0.10 to 0.09) indicating MAR, ECOG shows no difference in either unadjusted or adjusted model (logHR -0.06, -0.16 to 0.03) indicating MCAR, and PD-L1 exhibits differences in both models indicating MNAR.
 
@@
 The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**).
-By cross-checking all resulting diagnostic parameters to expected estimates (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness mechanisms and thereby help elucidate if analytic approaches such as imputation analyses are viable options.
+By cross-checking all resulting diagnostic parameters to expected estimates as illustrated in in the above examples (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness mechanisms and thereby help elucidate if analytic approaches such as imputation analyses are viable options.
 
-```{=html}
-<!--
+The `smdi_style_gt()` function is an ancillary function that takes an object of class `smdi` and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.
 
-The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.
-
--->
-```
 # Discussion

### 2023-11-21 manuscript/supplement.qmd
finished reviewer #2 review and finalized all responses, supplement and manuscript updates.Sending revised version now to  co-author review.

@@
 The below directed acyclic diagrams (DAG) [@Moreno-Betancur2018; @sondhi2023a], where a directed edge represents a causal effect of one variable on another variable, outline and illustrate these assumptions along with a brief explanations and exemplary real-world EHR scenarios.
-Since the `smdi` package was developed with focus on partially observed confounders (as opposed to missingness in exposure or outcome), we simplified the below DAGs by leaving out the nodes for exposure and outcome.
+Since the `smdi` package was developed with focus on partially observed confounders (as opposed to missingness in exposure or outcome), we simplified the below DAGs by leaving out the nodes for exposure and outcome for improved readability.
 
@@
 The missingness (M) of the partially observed confounder (POC) can be explained by one or multiple fully observed confounders (C).
-That is, there is an edge between C to M and the missingness is at random (MAR).
+That is, there is an edge from C to M and the missingness is at random (MAR).
 
@@
 
-[Real-world example:]{.underline} Patients with a certain performance status, which not observed in the dataset, are more likely to receive a lab test.
+[Real-world example:]{.underline} Patients with a certain performance status, which is not observed in the dataset, are more likely to receive a lab test.
 
@@
 
-This sections provides extensive information and details on key statistical principles and methods employed in the `smdi` package.
-Details on all package functions can also be found in the corresponding documentation section of the `smdi` package website
-
-<https://janickweberpals.gitlab-pages.partners.org/smdi/reference/index.html>
-
-or accessed in `R` by executing the the function name preceded with a question mark, e.g.:
+This section provides extensive information and details on key statistical principles and methods employed in the `smdi` package.
+Details on all package functions can also be found in the corresponding [documentation section of the `smdi` package website](https://janickweberpals.gitlab-pages.partners.org/smdi/reference/index.html) or accessed in `R` by executing the function name preceded with a question mark, e.g.:
 
@@
 
+A comprehensive illustration of an exemplary workflow of the `smdi` package can be found in the vignette [**`Routine structural missing data diagnostics`**](https://janickweberpals.gitlab-pages.partners.org/smdi/articles/b_routine_diagnostics.html) on the package website.
+
+### Utility functions
+
+The `smdi` package comes with a few in-built utility functions which are called in the background by other `smdi` functions, but which can also be used as standalone functions by users if desired.
+
+`smdi_check_covar()` checks for any partially observed covariate in a given dataset (specified by the `data` parameter) and returns a vector with the names of covariates exhibiting missingness.
+With the `covar` parameter, users can specify any specific covariates(s) of interest or, in case this parameter is not specified, any covariate with at least one missing value is returned.
+The function also provides further logical checks, e.g., if all covariates specified in `covar` are actually present in the dataset and if all ovariates specified in `covar` are truly missing (as sometimes missingness is explicitly coded as a "Missing" level or similar instead of `NA`).
+
+`smdi_na_indicator()`: This function takes a dataset (`data` parameter) and creates binary missing indicator variables for covariates with at least one missing value.
+If `covar` is not specified, all possible variables in the dataset are considered and if `covar` is specified with covariates(s) of interest, only those are considered.
+With the logical `drop_NA_col` parameter, users can decide if the original `covar` variables should be dropped or retained in the dataset in addition to the created binary missing indicator variables.
+
+`smdi_style_gt()` takes an object of class `smdi` (i.e., the output of the `smdi_diagnose()` function which returns all three group diagnostics) and formats the table as a [publication-ready gt() table object](https://janickweberpals.gitlab-pages.partners.org/smdi/articles/b_routine_diagnostics.html#publication-ready-gt-style-table) [@gt] which can be [exported to any format](https://janickweberpals.gitlab-pages.partners.org/smdi/articles/b_routine_diagnostics.html#smdi-table-export) that is supported by `gt::gtsave()` including .pdf or .docx files.
+
+### Descriptives
+
+`smdi_summarize()` and `smdi_vis()` [compute and visualize the number and proportion of missing values](https://janickweberpals.gitlab-pages.partners.org/smdi/articles/b_routine_diagnostics.html#missingness-proportions) for every covariate specified with `covar`.
+The output is a simple table (`smdi_summarize())` or a vertical bar chart plot (`ggplot2` object [@ggplot2], `smdi_vis()`).
+Optionally, results can also be returned stratified by another variable which can be specified using the `strata` parameter.
+
+`gg_miss_upset()` and `md.pattern()` are re-exports of the naniar [@naniar] and mice [@mice] packages, respectively, both of which are useful functions to inspect missing data patterns.
+The `gg_miss_upset()` achieves this by using set visualization techniques as described in detail by Ruddle et al.
+[@ruddle2022].
+In brief, the function returns a bar chart (*Upset plot*) that visualizes the number of intersecting missing observations across all POCs.
+It can be thought of as a high-dimensional Venn diagram for which, in case of monotone patterns, all or the majority of combinations of cells are missing simultaneously.
+The `md.pattern()` function has a similar functionality and returns a matrix in which each row corresponds to an observed missing data pattern.
+The rows and columns of the matrix are sorted in increasing amounts of missing information and thereby give an overview on how frequent a certain missing data pattern is.
+
 ### Group 1 diagnostics
 
-The featured group 1 diagnostics in `smdi` focus on investigating potential differences in the distribution of characteristics between patients with and without an observed value for a partially observed confounder.
+The featured group 1 diagnostics in `smdi` focus on investigating potential differences in the distribution of characteristics between patients with and without an observed value for a POC.
 To that end, different analytic approaches are leveraged which are described in detail for each function below.
 
--   `smdi_hotelling()`: This function computes a two-sample Hotelling's T-squared test for the difference in two multivariate means based on the Hotelling's T-Square test statistic [@hotelling1931; @Hotelling].
+`smdi_hotelling()`: This function computes a two-sample Hotelling's T-squared test for the difference in two multivariate means based on the Hotelling's T-Square test statistic [@hotelling1931; @Hotelling].
 It assesses whether there is a statistically significant difference between the means of two groups in multivariate data.
@@
 
--   `smdi_little()`: Little's test, as opposed to Hotelling's test, performs a single global test to evaluate the missing completely at random (MCAR) assumption [@little1988].
-It was developed since with an approach like Hotelling's test, there can be concerns regarding multiplicity due to the multiple testing of each variable.
+`smdi_little()`: Little's test, as opposed to Hotelling's test, performs a single global test to evaluate the missing completely at random (MCAR) assumption [@little1988].
+It was developed since an approach like Hotelling's test can raise concerns regarding multiplicity due to the multiple testing of each variable.
 Little's test first [identifies a set of subgroups that share the same missing data patterns $j$](https://github.com/njtierney/naniar/blob/master/R/mcar-test.R#L54C4-L54C4).
@@
 
-For both `smdi_hotelling()` and `smdi_little()`, a high test statistic and a low p-value would indicate differences in the groups compared.
+For both `smdi_hotelling()` and `smdi_little()`, a high test statistic and a low p-value would indicate differences in the groups compared and reject the null hypothesis that the missingness follows a MCAR mechanism.
 A limitation of both Hotelling's and Little's test is that they assume continuous variables following multivariate normality.
@@
 
--   `smdi_asmd()`: The `smdi_asmd` functions computes the absolute standardized mean difference (ASMD) as a statistical measure for assessing dissimilarities in disease and patient characteristics between individuals with and without observed value for a specific partially observed confounder (POC) of interest.
+`smdi_asmd()`: The `smdi_asmd` function computes the absolute standardized mean difference (ASMD) as a statistical measure for assessing dissimilarities in disease and patient characteristics between individuals with and without observed value for a specific POC of interest.
 If the median/average ASMD is high, this may indicate imbalance in patient covariate distributions which may be indicative of the POC following a missing at random (MAR) mechanims, i.e.
@@
 
--   `smdi_rf()`: The Group 2 diagnostics, implemented in the `smdi_rf()` function, assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
-This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given exposure, outcome, follow-up time, and covariates plus missingness indicator for other POC as the predictors.
+`smdi_rf()`: The Group 2 diagnostics, implemented in the `smdi_rf()` function, assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function.
+This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given exposure, outcome, and covariates plus missingness indicator for other POC as the predictors.
 We chose a random forest classification model for the purpose of this package due its many beneficial features within the context of data structures that are frequently encountered in routine healthcare databases:
@@
 
-    -   Multiple other studies have also reported good result of this random forest-based approach [@sondhi2023a; @beaulieu-jones2018].
+-   Multiple other studies have previously reported good result of this random forest-based approach for the purpose of diagnosing missing data [@sondhi2023a; @beaulieu-jones2018].
 
-The default training parameters of the random forest model include a train-test split of the data (default if 70% training and 30% testing) which can be changed using the `train_test_ratio` parameter and the number of trees to grow using `ntree` (default is 1000 decision trees since a higher number of trees typically give more stable results).
+The default training parameters of the random forest model include a train-test split of the data (default if 70% training and 30% testing), which can be changed using the `train_test_ratio` parameter, and the number of trees to grow using `ntree` (default is 1000 decision trees since a higher number of trees typically give more stable results).
 Users can optionally select `tune = TRUE` which will perform a 5-fold cross validation and a random search for the optimal number of variables randomly sampled as candidates at each split (`mtry`).
@@
 
+-   `tune`: if TRUE, a 5-fold cross validation is performed combined with a random search for the optimal number of optimal number of variables randomly sampled as candidates at each split (`mtry`).
+FALSE is the default due to potentially extensive computation times.
+
 -   `set_seed`: seed for reproducibility, defaults to 42
@@
 
-To evaluate the association of the misssing indicator variable of a POC ($M_{POC}$) and the studied outcome in Group 3 Diagnostics, conventional univariate and multivariate regression models which are found used in the vast majority of real-world evidence studies are employed.
-The selection of regression type needs to be chosen using the `model` parameter and is determined by the type of outcome that is studied.
-The possible outcome models include linear regression [@stats] for continuous outcomes (@eq-linear), logistic regression [@glmnet] for binary outcomes (@eq-logistic) and a Cox proportional hazards regression [@survival] for time-to-event outcomes (@eq-cox) with $X_j$ indicating other covariates included in the multivariate models.
-The `form_lhs` specifies the left-hand side of the outcome formula, which, in case of `model = "linear"` or `model = "logistic` just reflects the name of the column that contains the outcome and the form `Surv(time, status)` for `model ="cox"` indicating the time-to-event variable and the event status (0/1).
-The output of the functions automatically returns results for both the univariate and the multivariate models and a user has the choice to have the $\beta_1M_{POC}$ estimate exponentiated or not using the logical `exponentiated` parameter.
+To evaluate the association of the misssing indicator variable of a POC ($M_{POC}$) and the studied outcome in Group 3 Diagnostics, conventional univariate and multivariate regression models are employed (latter of which are adjusted for other covariates $X_j$).
+The type of regression model needs to be chosen using the `model` parameter and is determined by the type of outcome that is being studied.
+Users can choose from multiple possible outcome models, including:
 
-$$Y = \beta_0 + \beta_1M_{POC} + ...
-+ \beta_2X_j$$ {#eq-linear} $$\ln(\frac{p}{p-1}) = \beta_0 + \beta_1M_{POC} + ...
-+ \beta_2X_j$$ {#eq-logistic} $$h(t) = h_0(t) e^{\sum{\beta_1M_{POC} + ...
-+  \beta_2X_j}} $$ {#eq-cox}
+-   Linear regression [@stats] for continuous outcomes (@eq-linear)
+
+-   Cox proportional hazards regression [@survival] for time-to-event outcomes (@eq-cox).
+
+-   Generalized linear models (e.g., logistic regression [@glmnet], @eq-logistic) for which the family of conditional distributions of the outcome variable can be selected using the `glm_family` parameter and which needs to be an object of type `stats::family`.
+Possible options are:
+
+    -   binomial(link = "logit")
+    -   gaussian(link = "identity")
+    -   Gamma(link = "inverse")
+    -   inverse.gaussian(link = "1/mu\^2")
+    -   poisson(link = "log")
+    -   quasi(link = "identity", variance = "constant")
+    -   quasibinomial(link = "logit")
+    -   quasipoisson(link = "log")
+
+The `form_lhs` specifies the left-hand side of the outcome formula, which, in case of `model = "linear"` or `model = "glm"`, just reflects the name of the column that contains the outcome and the form `Surv(time, status)` for `model = "cox"` indicating the time-to-event variable and the event status (0/1).
+The output of the functions automatically returns results for both univariate and multivariate models and a user has the choice to exponentiate the $\beta_1M_{POC}$ estimate if desired using the `exponentiated` parameter.
+
+$$Y = \beta_0 + \beta_1M_{POC} + ... + \beta_2X_j$$ {#eq-linear}
+
+$$h(t) = h_0(t) e^{\sum{\beta_1M_{POC} + ... +  \beta_2X_j}} $$ {#eq-cox}
+
+$$\ln(\frac{p}{p-1}) = \beta_0 + \beta_1M_{POC} + ... + \beta_2X_j$$ {#eq-logistic}
 

### 2023-12-05 manuscript/manuscript.qmd
implemented wording suggested by RJD co-author review and adjusted worde count

@@
 
-[**Word count:**]{.underline} 2,514 words
+[**Word count:**]{.underline} 2,522 words
 
@@
 
-The package does come with certain limitations, such that the true underlying mechanism causing the missing data can never be inferred with absolute certainty from the observed data.
-Therefore, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care and contextualize potential reasons for missingness.
+The package should be used with certain limitations in mind.
+Most importantly, the true underlying mechanism causing the missing data can never be inferred with absolute certainty from the observed data.
+Therefore, it is important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care, which could be system-specific, and contextualize potential reasons for missingness.
 This collaborative approach allows for a contextualized understanding of potential causes for missing data in EHR.

### 2023-12-05 manuscript/manuscript.qmd
last minute typo correction

@@
 Sub-figure c) illustrates absolute standardized mean differences (ASMD) in patient characteristics between patients with and without a value observed for the PD-L1 (pdl1_num) biomarker as a measure of imbalance.
-Sub-figure d) demonstrates the variable importance of fully observed covariates for predicting missingness in the partially oberserved ECOG performance score variable (ecog_cat)."
+Sub-figure d) demonstrates the variable importance of fully observed covariates for predicting missingness in the partially observed ECOG performance score variable (ecog_cat)."
 #| echo: false

### 2023-12-06 manuscript/manuscript.qmd
addressed editorial formatting requests: moved 4 required disclosure statements immediately before the References list, converted figures to PNG instead of jpeg, added Contributorship Statement

@@
 [**Keywords:**]{.underline} Missing data, Confounder, EHR, R, Software, Real-World Evidence
-
-[**Funding:**]{.underline} This project was supported by Master Agreement 75F40119D10037 from the US Food and Drug Administration (FDA).
-
-[**Disclosures/COI:**]{.underline} The FDA approved the study protocol, statistical analysis plan and reviewed and approved this manuscript.
-Coauthors from the FDA participated in the results interpretation and in the preparation and decision to submit the manuscript for publication.
-The FDA had no role in data collection, management, or analysis.
-The views expressed are those of the authors and not necessarily those of the US FDA.
-Janick Weberpals reports prior employment by Hoffmann-La Roche and previously held shares in Hoffmann-La Roche.
-Pamela Shaw is a named inventor on a patent licensed to Novartis by the University of Pennsylvania for an unrelated project.
-Sengwee Toh serves as a consultant for Pfizer, Inc.
-and TriNetX, LLC..
-Robert J Glynn has received research funding through his employer from Amarin, Kowa, Novartis, and Pfizer.
-Dr.
-Desai reports serving as Principal Investigator on investigator-initiated grants to the Brigham and Women's Hospital from Novartis, Vertex, and Bristol-Myers-Squibb on unrelated projects.
-All remaining authors report no disclosures or conflicts of interest.
-
-[**Analytical code and data sharing statement:**]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2.
-All R code, materials and depedencies can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript> or <https://github.com/janickweberpals/smdi-manuscript>.
-The R package presented in this study can be downloaded from the comprehensive R archive network (CRAN) via `install.packages("smdi")` (version 0.2.2 at time of manuscript submission) or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.
-
-[**Acknowledgments:**]{.underline} We would like to thank all beta testers and attendees of the Division of Pharmacoepidemiology and Pharmacoeconomics Methods Incubator who gave valuable feedback on early versions of the `smdi` R package.
 
@@
 
+[**Funding Statement:**]{.underline} This project was supported by Master Agreement 75F40119D10037 from the US Food and Drug Administration (FDA).
+
+[**Competing Interests Statement:**]{.underline} The FDA approved the study protocol, statistical analysis plan and reviewed and approved this manuscript.
+Coauthors from the FDA participated in the results interpretation and in the preparation and decision to submit the manuscript for publication.
+The FDA had no role in data collection, management, or analysis.
+The views expressed are those of the authors and not necessarily those of the US FDA.
+Janick Weberpals reports prior employment by Hoffmann-La Roche and previously held shares in Hoffmann-La Roche.
+Pamela Shaw is a named inventor on a patent licensed to Novartis by the University of Pennsylvania for an unrelated project.
+Sengwee Toh serves as a consultant for Pfizer, Inc.
+and TriNetX, LLC..
+Robert J Glynn has received research funding through his employer from Amarin, Kowa, Novartis, and Pfizer.
+Dr.
+Desai reports serving as Principal Investigator on investigator-initiated grants to the Brigham and Women's Hospital from Novartis, Vertex, and Bristol-Myers-Squibb on unrelated projects.
+All remaining authors report no disclosures or conflicts of interest.
+
+[**Data Availability Statement:**]{.underline} The R package presented in this study and corresponding data can be downloaded from the comprehensive R archive network (CRAN) via `install.packages("smdi")` (version 0.2.2 at time of manuscript submission) or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.
+This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2.
+All R code, materials and dependencies can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript> or <https://github.com/janickweberpals/smdi-manuscript>.
+
+[**Contributorship Statement:**]{.underline} JW designed and developed the `smdi` R package and drafted the manuscript.
+SRR, PAS, HL, BGH, ST, JGC, KJD, FT, WL, JL, JJH, RJG, RJD contributed to the conception, design and interpretation and provided important feedback.
+All authors critically reviewed the manuscript for important intellectual content and approved of the final version of the manuscript.
+
+[**Acknowledgments:**]{.underline} We would like to thank all beta testers and attendees of the Division of Pharmacoepidemiology and Pharmacoeconomics Methods Incubator who gave valuable feedback on early versions of the `smdi` R package.
+
+{{< pagebreak >}}
+
 # References {.unnumbered}
@@
 
-knitr::include_graphics(here::here("figures", "Figure_1_workflow.jpg"))
+knitr::include_graphics(here::here("figures", "Figure_1_workflow.png"))
 ```
@@
 
-jpeg(here::here("figures", "Figure_2_visuals.jpg"), width = 22, height = 11, units = "in", res=400)
+png(here::here("figures", "Figure_2_visuals.png"), width = 22, height = 11, units = "in", res=400)
 panel
@@
 #| fig-height: 11
-knitr::include_graphics(here::here("figures", "Figure_3_results.jpg"))
+knitr::include_graphics(here::here("figures", "Figure_3_results.png"))
 ```

### 2024-01-02 manuscript/manuscript.qmd
R2 addressing last minor edit requested by reviewer #1: opening statement on RWE using claims and EHR and linkages; also: upgraded to R 4.3.2

@@
 
-Administrative health insurance claims databases have traditionally been the foundation for pharmacoepidemiologic studies.
-However, a notable drawback lies in their inability to capture detailed clinical nuances.
+Administrative health insurance claims databases and electronic health records (EHR) are important data modalities to generate real-world evidence (RWE) if they are found fit-for-purpose for the study question at hand.
+While administrative health insurance claims databases have traditionally been the backbone for the majority of pharmacoepidemiologic studies, a notable drawback lies in their inability to capture important clinical prognostic factors like vital signs and labs.
 To overcome this limitation, substantial initiatives are underway, for instance in the FDA Sentinel initiative [@Desai2021], linking claims databases and electronic health records (EHR) to generate real-world evidence (RWE) and complement data from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021].
-Due to their capture of clinical details, EHR can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022].
+Due to their capture of clinical details, EHR can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups [@asfaw2022].
 At the moment, substantial efforts focusing on the linkage of claims databases and EHR are underway, for instance, in the FDA Sentinel Initiative [@Desai2021].

### 2024-01-09 manuscript/manuscript.qmd
final JAMIO-2023-0165.R2 as resubmitted to the journal; addressed minor comment in opening statement

@@
 
-[**Word count:**]{.underline} 2,522 words
+[**Word count:**]{.underline} 2,554 words
 
@@
 
-Administrative health insurance claims databases and electronic health records (EHR) are important data modalities to generate real-world evidence (RWE) if they are found fit-for-purpose for the study question at hand.
+Administrative health insurance claims databases and electronic health records (EHR) are important data sources to generate real-world evidence (RWE) when they are found fit-for-purpose for the study question at hand.
 While administrative health insurance claims databases have traditionally been the backbone for the majority of pharmacoepidemiologic studies, a notable drawback lies in their inability to capture important clinical prognostic factors like vital signs and labs.
